---
title: "Blood Cost modeling"
subtitle: "A brief introduction"
date: "2025-05-16"
author:
  - name: "Harmon Bhasin"
    affiliation: "SecureBio"
format:
  html:
    toc: true
    code-fold: true
    toc-depth: 2
    number-sections: true
    theme: cosmo
  pdf:
    toc: true
---

*NB: I tried to model the writing of this similar to one of Charlie Whittaker's latest papers, https://www.sciencedirect.com/science/article/pii/S1755436524000732.*

# Methods

We develop a cost modeling framework to estimate the annual expense of operating a blood-based biosurveillance system in the United States designed to detect the emergence of an HIV-like pathogen. For this analysis, we define detection as the moment when the number of sequencing reads attributable to this pathogen surpasses a predefined threshold. By referring to the pathogen as "HIV-like," we mean it exhibits similar epidemiological characteristics to HIV, including growth rate, shedding patterns, and a constant shedding rate derived from existing untargeted metagenomic sequencing data on people with HIV.

Accurately estimating surveillance costs necessitates modeling potential pathogen outbreaks. To evaluate surveillance requirements, we model a standardized outbreak and then determine the sequencing depth needed to ensure detection before the epidemic reaches various cumulative incidence thresholds. We model different pooling strategies by varying the size of the pools, specifically scenarios in which all individual samples received in a given week—each sample representing a single individual's donation—are combined into one large sequencing pool. This approach allows us to determine the optimal sequencing depth required to reliably detect the pathogen, directly informing our overall cost estimates.

## Outbreak Modeling via Exponential Growth

Viral outbreaks are commonly modeled using a simple deterministic exponential growth framework, a standard practice for initial outbreak simulations (NB: insert appropriate citation). We simulate the number of new infections in each week, $t$, using the exponential growth model $I(t) = I_1 \cdot e^{r(t-1)}$, where $r$ is the pathogen's exponential growth rate, and $I_1$ is the initial number of infections in the first week of the simulation. To ensure a realistic outbreak scenario, we set $I_1 = 100$, which provides a sufficient initial seed for the epidemic and avoids premature extinction (NB: Dan said to set this high enough such that an outbreak actually happens, probably should find a citation for this).

We employ a pre-defined growth rate drawn from a previous epidemiological analysis of HIV (NB: https://www.nature.com/articles/nature19827). We also set the total susceptible population, $N$, to approximate the current US population size of approximately 300 million. Given the fixed growth rate, we can calculate the number of weeks required for the cumulative number of infections to reach any given threshold ($X\%$ of the total population).

Importantly, reaching the specified cumulative incidence threshold in the model does not necessarily imply real-world detection at that exact time point. Actual detection might occur with a lag, meaning the outbreak could continue unnoticed beyond this theoretical threshold. Nevertheless, this initial estimate provides a critical reference point, guiding the approximate simulation duration required for further modeling steps.

## Shedding Dynamics and Population

To simplify our modeling approach, we assume each infected individual sheds the pathogen continuously at a constant rate over a 12-week period following infection (NB: cite study). Although actual HIV shedding varies considerably over time, our model focuses specifically on the acute infection phase, characterized by high and relatively consistent shedding, after which viral loads typically decline before eventually increasing again with progression to AIDS.

This constant shedding rate, denoted as $\mu$, is derived from existing data on similar viral pathogens. Thus, the number of individuals actively shedding in a given week $t$, denoted $S(t)$, consists of all individuals infected during the previous 12 weeks:

$S(t) = \sum_{k=\max(1, t-11)}^{t} I(k).$

We estimate the expected relative abundance of our modeled HIV-like pathogen in blood per infected person (the shedding rate $\mu$), using existing metagenomic sequencing data from individuals early in their HIV infection. The rationale behind focusing on sequencing data from people early in their HIV infection is because HIV's viremia is quite variable during the infection period, but is well known to shed quite a lot during the initial weeks of infections; hence, we wanted to capture this pattern (NB: improve explanation). 

## Sampling and Read Generation

We assume sampling is independent across weeks, meaning that an individual's probability of being selected in one week does not affect their chance of selection in subsequent weeks. However, we acknowledge that actual donation frequency depends on the sample type—for example, whole blood donations typically occur at most once every 8 weeks, whereas plasma donations can be made up to twice per week. Despite this variation, we maintain the independence assumption to simplify modeling, as the practical impact is minimal given the large donor population size relative to our weekly sample.

Each week, we randomly select a sample of size $P$, which represents our pool size, from the total population $N$. Given that $N$ is large, sampling is performed with replacement, ensuring each individual has an equal and independent probability of being selected.

The number of actively shedding individuals detected in our weekly sample is modeled by a binomial distribution, $\text{Sampled shedders} \sim \text{Binomial}\left(P, \frac{S(t)}{N}\right)$, where $S(t)$ represents the number of individuals actively shedding the pathogen in week $t$.

From this sampled group, we calculate the expected number of pathogen reads detected each week as, $E[R(t)] = \frac{\text{Sampled shedders}}{P} \times \mu \times D$. Here, $\mu$ denotes the shedding rate—the expected relative abundance of pathogen per infected individual—and $D$ is the sequencing depth.

To accommodate the inherent variability and overdispersion typically observed in pathogen sequencing data, the actual pathogen reads detected each week $t$ are modeled using a negative binomial distribution, $R(t) \sim \text{NegativeBinomial}(r, p)$, where the parameters $r$ and $p$ are chosen such that the distribution has mean $E[R(t)]$ and a variance determined by an overdispersion parameter $\alpha$, which we estimate directly from empirical data. This modeling approach robustly captures the stochastic nature and realistic variability in pathogen detection during surveillance.

## Detection Probability via Monte Carlo Simulations

Due to the inherent randomness and complexity of infectious disease dynamics, we use Monte Carlo simulations to capture the variability in epidemic trajectories and detection outcomes. Monte Carlo simulations involve repeatedly running computational models that advance the epidemic week by week, applying probabilistic rules to reflect both the uncertainty and natural variability in pathogen spread and detection.

In our approach, each simulated epidemic progresses weekly, with stochastic sampling methods used to reflect the uncertainty inherent in detecting pathogen reads within a large population. Simulations continue until cumulative pathogen reads surpass a predefined detection threshold: $\sum_{t=1}^{T_{detect}} R(t) \geq \theta$. Upon reaching this threshold, we record the cumulative incidence of infections, quantifying the epidemic scale precisely at the point of detection.

By conducting numerous Monte Carlo simulations, we generate a robust distribution of potential detection incidences. This distribution allows us to compute the probability of detecting the pathogen before reaching a specified cumulative incidence threshold (e.g., X%), as well as the median incidence at detection and the interquartile range (IQR). Thus, we comprehensively characterize potential scenarios of early versus late detection.

Finally, we identify the optimal weekly sequencing depth $D^*$ through a binary search method. This procedure determines the minimal sequencing depth necessary to achieve our targeted detection probability, defined as:

$D^* = \min\{D : P(\text{detect before } X\% \text{ incidence}) \geq p_{target}\}$

## Cost Modeling Framework

We estimate the annual surveillance cost by decomposing weekly expenditures into sequencing-related and sample collection components. Each week, the cost includes the product of sequencing depth $D$ and the combined cost per read for sequencing and processing ($C_{\text{seq}} + C_{\text{proc}}$), as well as the product of the number of samples $P$ and the per-sample collection cost $C_{\text{samp}}$. The annual cost $C_{\text{annual}}$ is then calculated by multiplying this weekly cost by 52 weeks:

$C_{\text{annual}} = [(D \times (C_{\text{seq}} + C_{\text{proc}})) + (C_{\text{samp}} \times P)] \times 52$

In this model, we define the variables as follows: sequencing depth $D$ represents the number of reads required weekly to detect the pathogen reliably by the time the pathogen hits $X\%$ cumulative incidence, $C_{\text{seq}}$ is the sequencing cost per read specific to the chosen sequencing platform, and $C_{\text{proc}}$ is the per-read cost associated with processing (this is a simplification a), including sample preparation and library preparation. Additionally, $C_{\text{samp}}$ denotes the cost of collecting and shipping a single sample, and $P$ is the total number of samples processed each week (NB: which as of now, is just the number of individuals in the pool; this may change if we instead have samples come pre-pooled).

Our cost analysis relies on several key assumptions: sequencing depth $D$ is optimized according to detection probability thresholds; sequencing and processing costs per read remain constant and consistent across all samples; per-sample collection costs encompass all logistics and remain uniform; and weekly sampling and sequencing processes do not vary throughout the year.

# Results

*Note: This text is going to be rougher than the text above.*

We analyze data from Chen et al. ([https://doi.org/10.1128/jvi.00804-18](https://doi.org/10.1128/jvi.00804-18)), a study that examined HIV-1 genetic diversity in blood and the female genital tract (FGT) of three women from a prospective cohort in South Africa, collected within two weeks following initial detection of infection. While the study provides HIV sequencing reads from plasma, the total sequencing depth per sample is reported only as a range ("40 million to 250 million reads"). We therefore use the geometric mean of this range (approximately 100 million reads) as our sequencing depth estimate. For quantifying HIV reads per individual, we use data from the final available sample for each participant, all obtained within an 11-day period.



```{r}
#| label: load-packages
#| output: false

library(tidyverse)
library(furrr)
library(scales)
library(future)
library(future.apply)
library(dplyr)

```

```{r}
#| label: aux-functions

# Function to solve for T given r, I1, and X
solve_time_period <- function(r, I1, X_pct, N) {
  target_total <- (X_pct / 100) * N
  if (r == 0) {
    # Constant infections case
    return(target_total / I1)
  } else {
    # Exponential growth case
    # T = (1/r) * ln(1 + (X*N*(e^r - 1))/(100*I1))
    exp_r <- exp(r)
    ratio <- (X_pct * N * (exp_r - 1)) / (100 * I1)
    T <- (1/r) * log(1 + ratio)
    return(T)
  }
}

# Function to calculate infections at time t
calc_infections <- function(t, I1, r) {
  I1 * exp(r * (t - 1))
}

# Function to calculate shedding population at time t
calc_shedding <- function(t, I1, r, shedding_weeks) {
  start_week <- max(1, t - shedding_weeks + 1)
  end_week <- t
  weeks <- start_week:end_week
  infections <- sapply(weeks, function(w) calc_infections(w, I1, r))
  sum(infections)
}

# Function to calculate cumulative incidence as percentage of population
calc_cumulative_incidence <- function(week, I1, r, N) {
  # Calculate total infections from week 1 to target week
  if (r == 0) {
    total_infections <- I1 * week
  } else {
    total_infections <- I1 * (exp(r * week) - 1) / (exp(r) - 1)
  }
  # Return as percentage of population
  return(100 * total_infections / N)
}

```

```{r}
#| label: simulate-detection-distribution

# Function to simulate detection timing distribution
simulate_detection_distribution <- function(r, I1, P, D, theta, mew, alpha, shedding_weeks, N, max_weeks = 1000) {
  # Run epidemic week by week until detection
  week <- 1
  total_reads <- 0
  while (week <= max_weeks && total_reads < theta) {
    # Calculate shedding population for current week
    S_t <- calc_shedding(week, I1, r, shedding_weeks)
    S_t <- round(S_t)
    S_t <- min(S_t, N)  # Cap at population size
    if (S_t > 0 && P < N) {
      # For large populations, approximate with binomial
      sampled_shedders <- rbinom(1, P, S_t / N)
      # Generate reads from negative binomial distribution
      expected_reads <- sampled_shedders / P * mew * D
      weekly_reads <- rnbinom(1, size = 1/alpha, mu = expected_reads)
      total_reads <- total_reads + weekly_reads
      if (total_reads >= theta) {
        # Detection occurred!
        cumulative_incidence <- calc_cumulative_incidence(week, I1, r, N)
        return(list(
          detected = TRUE,
          detection_week = week,
          cumulative_incidence = cumulative_incidence,
          total_reads = total_reads
        ))
      }
    }
    week <- week + 1
  }
  # No detection within max_weeks
  return(list(
    detected = FALSE, 
    detection_week = NA, 
    cumulative_incidence = NA,
    total_reads = total_reads
  ))
}

```

```{r}
#| label: analyze-detection-distribution

# Function to analyze detection distribution across many simulations
analyze_detection_distribution <- function(r, I1, P, D, theta, mew, alpha, shedding_weeks, N, n_sims = 1000) {
  # Run many simulations
  results <- replicate(n_sims, {
    simulate_detection_distribution(r, I1, P, D, theta, mew, alpha, shedding_weeks, N)
  }, simplify = FALSE)
  # Extract detection data
  detected_sims <- results[sapply(results, function(x) x$detected)]
  if (length(detected_sims) == 0) {
    return(list(detection_rate = 0, summary_stats = NA))
  }
  # Extract cumulative incidence at detection
  detection_incidences <- sapply(detected_sims, function(x) x$cumulative_incidence)
  # Calculate summary statistics
  summary_stats <- list(
    detection_rate = length(detected_sims) / n_sims,
    mean_incidence = mean(detection_incidences),
    median_incidence = median(detection_incidences),
    q25_incidence = quantile(detection_incidences, 0.25),
    q75_incidence = quantile(detection_incidences, 0.75),
    min_incidence = min(detection_incidences),
    max_incidence = max(detection_incidences)
  )
  return(list(
    detection_rate = summary_stats$detection_rate,
    summary_stats = summary_stats,
    raw_data = detection_incidences
  ))
}

```

```{r}
#| label: calculate-detection-probability

# Function to calculate detection probability from saved results
calc_detection_probability <- function(detection_results, threshold_incidence) {
  # Calculate P(detect | X% incidence) from saved results
  # detection_results: output from analyze_detection_distribution()
  # threshold_incidence: X% incidence threshold
  if (detection_results$detection_rate == 0) {
    return(0)
  }
  # Count simulations that detected at or before threshold
  detected_early <- sum(detection_results$raw_data <= threshold_incidence)
  total_sims <- length(detection_results$raw_data) / detection_results$detection_rate
  return(detected_early / total_sims)
}

# Function to find optimal sequencing depth using binary search
find_optimal_depth <- function(target_incidence, target_prob = 0.95, r, I1, P, theta, mew, alpha,
                              shedding_weeks, N, n_sims = 500, tolerance = 1e4) {
  cat(sprintf("Finding optimal depth for %.1f%% cumulative incidence at %.0f%% detection probability\n", 
              target_incidence, target_prob * 100))
  # Binary search bounds
  low <- 1e4    # Start with very low depth
  high <- 1e12   # Maximum reasonable depth
  # Track results for visualization
  tested_depths <- c()
  tested_probs <- c()
  while (high - low > tolerance) {
    mid <- round((low + high) / 2)
    cat(sprintf("Testing depth %.2e... ", mid))
    # Run simulations at this depth
    results <- analyze_detection_distribution(r, I1, P, mid, theta, mew, alpha, shedding_weeks, N, n_sims)
    p_detect <- calc_detection_probability(results, target_incidence)
    cat(sprintf("P(detect) = %.2f%%\n", p_detect * 100))
    # Store for visualization
    tested_depths <- c(tested_depths, mid)
    tested_probs <- c(tested_probs, p_detect)
    if (p_detect >= target_prob) {
      high <- mid  # Can use less sequencing
    } else {
      low <- mid   # Need more sequencing
    }
  }
  # Run final simulation at selected depth to confirm
  final_results <- analyze_detection_distribution(r, I1, P, high, theta, mew, alpha, shedding_weeks, N, n_sims * 2)
  final_p_detect <- calc_detection_probability(final_results, target_incidence)
  cat(sprintf("\nOptimal depth: %.2e reads/week\n", high))
  cat(sprintf("Final P(detect | %.1f%% incidence) = %.2f%%\n", target_incidence, final_p_detect * 100))
  return(list(optimal_depth = high, final_prob = final_p_detect, tested_depths = tested_depths, tested_probs = tested_probs, final_results = final_results))
}

```



```{r}
#| label: setup-functions

find_total_cost <- function(seq_depth, cost_of_seq, cost_of_proc, cost_of_sample, num_samples) {
  return(seq_depth * (cost_of_seq + cost_of_proc) + cost_of_sample * num_samples)
}

```



```{r}
#| label: load-data

# load in data
data <- read_csv("../data/new-hiv-data.csv", show_col_types = FALSE)

our_data <- data %>% 
  mutate(hiv_ra = hiv_reads / read_depth)

our_data

# Find the mew factor by taking the average of hiv ra over all HIV people.
#mew <- our_data %>% summarize(mean = mean(hiv_ra)) %>% pull(mean)

mew <- our_data %>% 
  summarize(geo_mean = exp(mean(log(hiv_ra), na.rm = TRUE))) %>% 
  pull(geo_mean)

#Best Approach: Coefficient of Variation Squared
#
#I recommend calculating overdispersion using the coefficient of variation squared (CV²) of the relative abundances. Here's why:
#
#1. Scale-free consistency: Both geometric mean and CV² are scale-free measures, making them compatible
#2. Biological interpretation: CV² represents the relative variability in viral shedding between individuals
#3. Mathematical alignment: For negative binomial, Variance = μ + α·μ², so CV² = Var/μ² ≈ α (when μ is small)
#4. Log-normal connection: Using geometric mean implies log-normal-like behavior; CV² naturally captures this variability
#
#Alternative Implementation

# Calculate overdispersion using CV² of relative abundances
cv_squared <- our_data %>%
  summarize(
    mean_ra = mean(hiv_ra, na.rm = TRUE),
    sd_ra = sd(hiv_ra, na.rm = TRUE),
    cv_squared = (sd_ra / mean_ra)^2
  ) %>%
  pull(cv_squared)

alpha <- cv_squared  # Use CV² directly as overdispersion parameter



```

From the three subjects in this study, we calculated a shedding rate ($\mu$) of $5.59 \times 10^{-5}$ using the geometric mean of the observed values. Additionally, we estimated the overdispersion parameter ($\alpha$) to be $0.69$, reflecting the observed variability in pathogen sequencing counts.

Using this shedding rate, we simulated pathogen detection scenarios under the following assumptions: a total population of 300 million individuals, a detection threshold of 100 sequencing reads, an initial infected population of 100 individuals, and a pathogen growth rate of 0.01336 derived from an existing study in *Nature* ([https://www.nature.com/articles/nature19827](https://www.nature.com/articles/nature19827)). We performed these simulations while systematically varying the pool size and cumulative incidence thresholds to assess their impact on surveillance efficiency.
 
```{r}
#| label: run-analysis-temp

# Run analysis to get detection distribution
#cat("Analyzing detection distribution...\n")
#results <- analyze_detection_distribution(r, I1, P, D, theta, mew, shedding_weeks, N, n_sims)
#
### Display summary statistics
#cat("\nDetection Distribution Summary:\n")
#cat(sprintf("Detection rate: %.2f%%\n", results$detection_rate * 100))
#if (results$detection_rate > 0) {
#  cat(sprintf("Mean cumulative incidence at detection: %.2f%%\n", results$summary_stats$mean_incidence))
#  cat(sprintf("Median cumulative incidence at detection: %.2f%%\n", results$summary_stats$median_incidence))
#  cat(sprintf("IQR: %.2f%% - %.2f%%\n", results$summary_stats$q25_incidence, results$summary_stats$q75_incidence))
#  cat(sprintf("Range: %.2f%% - %.2f%%\n", results$summary_stats$min_incidence, results$summary_stats$max_incidence))
#}
#
### Calculate detection probabilities at various cumulative incidence thresholds
#thresholds <- c(1.0, 5.0, 10.0)  # % cumulative incidence
#cat("\n\nDetection Probabilities at Different Cumulative Incidence Levels:\n")
#for (threshold in thresholds) {
#  p_detect <- calc_detection_probability(results, threshold)
#  cat(sprintf("P(detect | %.1f%% cumulative incidence) = %.2f%%\n", threshold, p_detect * 100))
#}
```

```{r}
#| label: find-seq-depth
#| fig-height: 10
#| fig-width: 8
#| fig-cap: "Figure 1: Total sequencing depth required per week to detect the pathogen with varied pool sizes and cumulative incidence."

## Model parameters
# CONSTANTS
N <- 300e6  # US population
theta <- 100  # detection threshold (reads)
shedding_weeks <- 12
r <- 0.01336      # weekly growth based on HIV
I1 <- 100     # 100 initial infections

# Find optimal sequencing depths for different target incidences
target_incidences <- c(0.1, 0.5, 1, 2.5, 5, 7.5, 10)  # Different cumulative incidence targets
pool_sizes <- c(16, 32, 64, 128, 256, 512, 1024, 2048)  # Different cumulative incidence targets
optimal_results <- list()


# Parallel
workers <- max(1, parallel::detectCores() - 1)
plan(multisession, workers = workers)

param_grid <- expand.grid(
  target_incidence = target_incidences,
  pool_size       = pool_sizes,
  stringsAsFactors = FALSE
)

#optimal_results <- future_apply(
#  X   = param_grid,          # each row is one job
#  MARGIN = 1,
#  FUN = function(row) {
#    find_optimal_depth(
#      target_incidence = row["target_incidence"],
#      target_prob      = 0.95,
#      r = r, I1 = I1, P = row["pool_size"],
#      theta = theta, mew = mew, alpha = alpha,
#      shedding_weeks = shedding_weeks,
#      N = N, n_sims = 500
#    )
#  },
#  future.seed = TRUE         # reproducible RNG streams
#)

# After the future_apply() call finishes
#results_dir <- "../results"        # adjust as needed
#if (!dir.exists(results_dir)) dir.create(results_dir)

#saveRDS(optimal_results,
#       file = file.path(results_dir, "optimal_results.rds"))

optimal_results <- readRDS("../results/optimal_results.rds")

#for (i in target_incidences) {
#for (j in pool_sizes) {
#  cat(sprintf("\n--- Target: %.1f%% cumulative incidence ---\n", target_incidences[i]))
#  optimal_results[[i]] <- find_optimal_depth(
#    target_incidence = target_incidences[i],
#    target_prob = 0.95,
#    r = r, I1 = I1, P = j, theta = theta, 
#    mew = mew, alpha = alpha, shedding_weeks = shedding_weeks, 
#    N = N, n_sims = 500
#  )
#}
#}

optimal_summary <- param_grid %>%         # nrow = length(optimal_results)
  mutate(
    optimal_depth = map_dbl(optimal_results, "optimal_depth"),
    achieved_prob = map_dbl(optimal_results, "final_prob"),
    q25_incidence = map_dbl(optimal_results,
                            c("final_results", "summary_stats", "q25_incidence")),
    q75_incidence = map_dbl(optimal_results,
                            c("final_results", "summary_stats", "q75_incidence")),
    annual_reads  = optimal_depth * 52
  )

optimal_summary <- as_tibble(optimal_summary)


ggplot(optimal_summary,
       aes(x = target_incidence,
           y = optimal_depth,
           colour = factor(pool_size),
           group  = pool_size)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = c(0.1, 0.5, 1, 2.5, 5, 7.5, 10)) +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
  scale_colour_brewer(palette = "Accent", name = "Pool size") +
  labs(title = "",
       y = "Sequencing depth per week to detect HIV-like pathogen",
       x = "Cumulative incidence (%)",
       colour = "Pool size") +
  theme_bw(base_size = 15) +
  theme(
    plot.title      = element_text(hjust = .5, face = "bold"),
    plot.subtitle   = element_text(hjust = .5),
    legend.position = "right"
  )


```

```{r}
#| label: cost-analysis
#| fig-height: 10
#| fig-width: 8
#| fig-cap: "Figure 2: Total cost required per year to detect the pathogen with varied pool sizes and cumulative incidence."

## Cost analysis for a specific depth
seq_c <- 2500 / 1e9
proc_c <- 2500 / 1e9
sample_c <- 100

# 1. add a new column with the cost for each depth
optimal_summary <- optimal_summary %>% 
  mutate(total_cost_per_year = find_total_cost(
           seq_depth      = optimal_depth,
           cost_of_seq    = seq_c,
           cost_of_proc   = proc_c,
           cost_of_sample = sample_c,
           num_samples    = pool_size
         )*52)

ggplot(optimal_summary,
       aes(x = target_incidence,
           y = total_cost_per_year,
           colour = factor(pool_size),
           group  = pool_size)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = c(0.1, 0.5, 1, 2.5, 5, 7.5, 10)) +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
  scale_colour_brewer(palette = "Accent", name = "Pool size") +
  labs(title = "",
       y = "Total cost per year to detect HIV-like pathogen ($)",
       x = "Cumulative incidence (%)",
       colour = "Pool size") +
  theme_bw(base_size = 15) +
  theme(
    plot.title      = element_text(hjust = .5, face = "bold"),
    plot.subtitle   = element_text(hjust = .5),
    legend.position = "right"
  )

#optimal_summary %>% select(target_incidence, pool_size, total_cost_per_year, q25_incidence,q75_incidence) %>%
#  filter(pool_size == 1024)

```

*NB: Upper limit on simulation was 1 trillion read pairs. I don't think we will use this many CI + pool size combinations, I was just interested. I would need to spend more time looking at these results, but I've listed some observations below.*

Observations about figure 1:

* Unless we have a large pool size (>= 512), detecting the pathogen below 1% CI seems not feasible
* Letting the pathogen hit a higher cumulative incidence means you can sequence less
* Once you get larger than 512 samples, the difference in seq depth seems to be quite minimal

Observations about figure 2 (NB: most likely will update this figure to only have sequencing + library prep cost, which we can make a slightly more complicated model for):

* The cost per sample dominates the total cost at higher pool sizes (>= 512)
* It seems like there isn't much money saved for waiting for a pathogen to hit higher cumultaive incidence, past 64 samples in a pool
* The most something like this would cost seems to be around 10 million dollars on the high end and ~350-500k on the low end

# @TODO 
- Get code checked.
