---
title: "Blood Cost modeling"
subtitle: "A brief introduction"
date: "2025-05-16"
author:
  - name: "Harmon Bhasin"
    affiliation: "SecureBio"
format:
  html:
    toc: true
    code-fold: true
    toc-depth: 2
    number-sections: true
    theme: cosmo
  pdf:
    toc: true
---

*NB: I tried to model the writing of this similar to one of Charlie Whittaker's latest papers, https://www.sciencedirect.com/science/article/pii/S1755436524000732. Should I include a table of the all the parameters similar to this paper?*

# Methods

We develop a cost modeling framework to estimate the annual expense of operating a blood-based biosurveillance system in the United States designed to detect the emergence of an HIV-like pathogen. For this analysis, we define detection as the moment when the number of sequencing reads attributable to this pathogen surpasses a predefined threshold. By referring to the pathogen as "HIV-like," we mean it exhibits similar epidemiological characteristics to HIV, including growth rate, shedding patterns, and a constant shedding rate derived from existing untargeted metagenomic sequencing data on people with HIV.

Accurately estimating surveillance costs necessitates modeling potential pathogen outbreaks. Thus, we simulate outbreaks under varying cumulative incidence levels and different pooling strategies, focusing on scenarios where all samples collected each week are combined into a single large pool for sequencing. This approach allows us to determine the optimal sequencing depth required to reliably detect the pathogen, directly informing our overall cost estimates.

## Outbreak Modeling via Exponential Growth

Viral outbreaks are commonly modeled using a simple deterministic exponential growth framework, a standard practice for initial outbreak simulations (NB: insert appropriate citation). We simulate the number of new infections in each week, $t$, using the exponential growth model $I(t) = I_1 \cdot e^{r(t-1)}$, where $r$ is the pathogen's exponential growth rate, and $I_1$ is the initial number of infections in the first week of the simulation. To ensure a realistic outbreak scenario, we set $I_1 = 100$, which provides a sufficient initial seed for the epidemic and avoids premature extinction (NB: Dan said to set this high enough such that an outbreak actually happens, probably should find a citation for this).

We employ a pre-defined growth rate drawn from a previous epidemiological analysis of HIV (NB: https://www.nature.com/articles/nature19827). We also set the total susceptible population, $N$, to approximate the current US population size of approximately 300 million. By varying the cumulative incidence threshold ($X\%$), this approach allows us to estimate roughly how many weeks would be needed to reach our desired outbreak size.

Importantly, reaching the specified cumulative incidence threshold in the model does not necessarily imply real-world detection at that exact time point. Actual detection might occur with a lag, meaning the outbreak could continue unnoticed beyond this theoretical threshold. Nevertheless, this initial estimate provides a critical reference point, guiding the approximate simulation duration required for further modeling steps.

## Shedding Dynamics and Population

To simplify our modeling approach, we assume that each infected individual sheds the pathogen continuously at the same constant rate over a period of 12 weeks following infection (NB: cite the study that lenni referred to that showed the 12 week major shedding). This constant shedding rate, denoted as $\mu$, is derived from existing data on similar viral pathogens. Thus, the number of individuals actively shedding in a given week $t$, denoted $S(t)$, consists of all individuals infected during the previous 12 weeks:

$S(t) = \sum_{k=\max(1, t-11)}^{t} I(k).$

We estimate the expected relative abundance of our modeled HIV-like pathogen per infected person (the shedding rate $\mu$), using existing metagenomic sequencing data from individuals early in their HIV infection. The rationale behind focusing on sequencing data from people early in their HIV infection is because HIV's viremia is quite variable during the infection period, but is well known to shed quite a lot during the initial weeks of infections, hence, we wanted to capture this pattern (NB: improve explanation). 

## Sampling and Read Generation

We assume that sampling is independent across weeks, meaning each individual's chance of being sampled in one week does not affect their probability of selection in subsequent weeks.

Each week, we randomly select a sample of size $P$, which represents our pool size, from the total population $N$. Given that $N$ is large, sampling is performed with replacement, ensuring each individual has an equal and independent probability of being selected.

The number of actively shedding individuals detected in our weekly sample is modeled by a binomial distribution, $\text{Sampled shedders} \sim \text{Binomial}\left(P, \frac{S(t)}{N}\right)$, where $S(t)$ represents the number of individuals actively shedding the pathogen in week $t$.

From this sampled group, we calculate the expected number of pathogen reads detected each week as, $E[R(t)] = \frac{\text{Sampled shedders}}{P} \times \mu \times D$. Here, $\mu$ denotes the shedding rate—the expected relative abundance of pathogen per infected individual—and $D$ is the sequencing depth.

The actual observed pathogen reads in any given week $t$ are assumed to follow a Poisson distribution around this expected value: $R(t) \sim \text{Poisson}\left(\frac{\text{Sampled shedders}}{P} \times \mu \times D\right)$. This framework allows us to robustly model and predict weekly pathogen read detections.


## Detection Probability via Monte Carlo Simulations

Due to the inherent randomness and complexity of infectious disease dynamics, we use Monte Carlo simulations to capture the variability in epidemic trajectories and detection outcomes. Monte Carlo simulations involve repeatedly running computational models that advance the epidemic week by week, applying probabilistic rules to reflect both the uncertainty and natural variability in pathogen spread and detection.

In our approach, each simulated epidemic progresses weekly, with stochastic sampling methods used to reflect the uncertainty inherent in detecting pathogen reads within a large population. Simulations continue until cumulative pathogen reads surpass a predefined detection threshold: $\sum_{t=1}^{T_{detect}} R(t) \geq \theta$. Upon reaching this threshold, we record the cumulative incidence of infections, quantifying the epidemic scale precisely at the point of detection.

By conducting numerous Monte Carlo simulations, we generate a robust distribution of potential detection incidences. This distribution allows us to compute the probability of detecting the pathogen before reaching a specified cumulative incidence threshold (e.g., X%), as well as the median incidence at detection and the interquartile range (IQR). Thus, we comprehensively characterize potential scenarios of early versus late detection.

Finally, we identify the optimal weekly sequencing depth $D^*$ through a binary search method. This procedure determines the minimal sequencing depth necessary to achieve our targeted detection probability, defined as:

$D^* = \min\{D : P(\text{detect before } X\% \text{ incidence}) \geq p_{target}\}$

## Cost Modeling Framework

We estimate the annual surveillance cost by decomposing weekly expenditures into sequencing-related and sample collection components. Each week, the cost includes the product of sequencing depth $D$ and the combined cost per read for sequencing and processing ($C_{\text{seq}} + C_{\text{proc}}$), as well as the product of the number of samples $P$ and the per-sample collection cost $C_{\text{samp}}$. The annual cost $C_{\text{annual}}$ is then calculated by multiplying this weekly cost by 52 weeks:

$C_{\text{annual}} = [(D \times (C_{\text{seq}} + C_{\text{proc}})) + (C_{\text{samp}} \times P)] \times 52$

In this model, we define the variables as follows: sequencing depth $D$ represents the number of reads required weekly to detect the pathogen reliably, $C_{\text{seq}}$ is the sequencing cost per read specific to the chosen sequencing platform, and $C_{\text{proc}}$ is the per-read cost associated with processing, including sample preparation and library preparation. Additionally, $C_{\text{samp}}$ denotes the cost of collecting and shipping a single sample, and $P$ is the total number of samples processed each week (also known, as a our pool size from earlier; NB: make sure don't double explain).

Our cost analysis relies on several key assumptions: sequencing depth $D$ is optimized according to detection probability thresholds; sequencing and processing costs per read remain constant and consistent across all samples; per-sample collection costs encompass all logistics and remain uniform; and weekly sampling and sequencing processes do not vary throughout the year.

# Results

*Note: This text is way more rough than the text above. This isn't even text, but really notes.*

We analyze the results from one study, https://doi.org/10.1128/jvi.00804-18, which "characterized HIV-1 diversity in the blood and female genital tract (FGT) within 2 weeks after detection of infection in three women enrolled in a unique prospective cohort in South Africa." While sequencing reads of HIV for plasma is given, sequnecing dpeth of the sample is given as a range, "40 million to 250 million", so we take the geometric mean to get a read depth of 100 million reads. We take the last sample from each person as the count of their HIV reads (which is all within at most an 11 day period).

```{r}
#| label: load-packages
#| output: false

library(tidyverse)
library(furrr)
library(scales)
library(future)
library(future.apply)
library(dplyr)

```

```{r}
#| label: aux-functions

# Function to solve for T given r, I1, and X
solve_time_period <- function(r, I1, X_pct, N) {
  target_total <- (X_pct / 100) * N
  if (r == 0) {
    # Constant infections case
    return(target_total / I1)
  } else {
    # Exponential growth case
    # T = (1/r) * ln(1 + (X*N*(e^r - 1))/(100*I1))
    exp_r <- exp(r)
    ratio <- (X_pct * N * (exp_r - 1)) / (100 * I1)
    T <- (1/r) * log(1 + ratio)
    return(T)
  }
}

# Function to calculate infections at time t
calc_infections <- function(t, I1, r) {
  I1 * exp(r * (t - 1))
}

# Function to calculate shedding population at time t
calc_shedding <- function(t, I1, r, shedding_weeks) {
  start_week <- max(1, t - shedding_weeks + 1)
  end_week <- t
  weeks <- start_week:end_week
  infections <- sapply(weeks, function(w) calc_infections(w, I1, r))
  sum(infections)
}

# Function to calculate cumulative incidence as percentage of population
calc_cumulative_incidence <- function(week, I1, r, N) {
  # Calculate total infections from week 1 to target week
  if (r == 0) {
    total_infections <- I1 * week
  } else {
    total_infections <- I1 * (exp(r * week) - 1) / (exp(r) - 1)
  }
  # Return as percentage of population
  return(100 * total_infections / N)
}

```

```{r}
#| label: simulate-detection-distribution

# Function to simulate detection timing distribution
simulate_detection_distribution <- function(r, I1, P, D, theta, mew, shedding_weeks, N, max_weeks = 1000) {
  # Run epidemic week by week until detection
  week <- 1
  total_reads <- 0
  while (week <= max_weeks && total_reads < theta) {
    # Calculate shedding population for current week
    S_t <- calc_shedding(week, I1, r, shedding_weeks)
    S_t <- round(S_t)
    S_t <- min(S_t, N)  # Cap at population size
    if (S_t > 0 && P < N) {
      # For large populations, approximate with binomial
      sampled_shedders <- rbinom(1, P, S_t / N)
      # Generate reads from Poisson distribution
      expected_reads <- sampled_shedders / P * mew * D
      weekly_reads <- rpois(1, expected_reads)
      total_reads <- total_reads + weekly_reads
      if (total_reads >= theta) {
        # Detection occurred!
        cumulative_incidence <- calc_cumulative_incidence(week, I1, r, N)
        return(list(
          detected = TRUE,
          detection_week = week,
          cumulative_incidence = cumulative_incidence,
          total_reads = total_reads
        ))
      }
    }
    week <- week + 1
  }
  # No detection within max_weeks
  return(list(
    detected = FALSE, 
    detection_week = NA, 
    cumulative_incidence = NA,
    total_reads = total_reads
  ))
}

```

```{r}
#| label: analyze-detection-distribution

# Function to analyze detection distribution across many simulations
analyze_detection_distribution <- function(r, I1, P, D, theta, mew, shedding_weeks, N, n_sims = 1000) {
  # Run many simulations
  results <- replicate(n_sims, {
    simulate_detection_distribution(r, I1, P, D, theta, mew, shedding_weeks, N)
  }, simplify = FALSE)
  # Extract detection data
  detected_sims <- results[sapply(results, function(x) x$detected)]
  if (length(detected_sims) == 0) {
    return(list(detection_rate = 0, summary_stats = NA))
  }
  # Extract cumulative incidence at detection
  detection_incidences <- sapply(detected_sims, function(x) x$cumulative_incidence)
  # Calculate summary statistics
  summary_stats <- list(
    detection_rate = length(detected_sims) / n_sims,
    mean_incidence = mean(detection_incidences),
    median_incidence = median(detection_incidences),
    q25_incidence = quantile(detection_incidences, 0.25),
    q75_incidence = quantile(detection_incidences, 0.75),
    min_incidence = min(detection_incidences),
    max_incidence = max(detection_incidences)
  )
  return(list(
    detection_rate = summary_stats$detection_rate,
    summary_stats = summary_stats,
    raw_data = detection_incidences
  ))
}

```

```{r}
#| label: calculate-detection-probability

# Function to calculate detection probability from saved results
calc_detection_probability <- function(detection_results, threshold_incidence) {
  # Calculate P(detect | X% incidence) from saved results
  # detection_results: output from analyze_detection_distribution()
  # threshold_incidence: X% incidence threshold
  if (detection_results$detection_rate == 0) {
    return(0)
  }
  # Count simulations that detected at or before threshold
  detected_early <- sum(detection_results$raw_data <= threshold_incidence)
  total_sims <- length(detection_results$raw_data) / detection_results$detection_rate
  return(detected_early / total_sims)
}

# Function to find optimal sequencing depth using binary search
find_optimal_depth <- function(target_incidence, target_prob = 0.95, r, I1, P, theta, mew, 
                              shedding_weeks, N, n_sims = 500, tolerance = 1e4) {
  cat(sprintf("Finding optimal depth for %.1f%% cumulative incidence at %.0f%% detection probability\n", 
              target_incidence, target_prob * 100))
  # Binary search bounds
  low <- 1e4    # Start with very low depth
  high <- 1e11   # Maximum reasonable depth
  # Track results for visualization
  tested_depths <- c()
  tested_probs <- c()
  while (high - low > tolerance) {
    mid <- round((low + high) / 2)
    cat(sprintf("Testing depth %.2e... ", mid))
    # Run simulations at this depth
    results <- analyze_detection_distribution(r, I1, P, mid, theta, mew, shedding_weeks, N, n_sims)
    p_detect <- calc_detection_probability(results, target_incidence)
    cat(sprintf("P(detect) = %.2f%%\n", p_detect * 100))
    # Store for visualization
    tested_depths <- c(tested_depths, mid)
    tested_probs <- c(tested_probs, p_detect)
    if (p_detect >= target_prob) {
      high <- mid  # Can use less sequencing
    } else {
      low <- mid   # Need more sequencing
    }
  }
  # Run final simulation at selected depth to confirm
  final_results <- analyze_detection_distribution(r, I1, P, high, theta, mew, shedding_weeks, N, n_sims * 2)
  final_p_detect <- calc_detection_probability(final_results, target_incidence)
  cat(sprintf("\nOptimal depth: %.2e reads/week\n", high))
  cat(sprintf("Final P(detect | %.1f%% incidence) = %.2f%%\n", target_incidence, final_p_detect * 100))
  return(list(optimal_depth = high, final_prob = final_p_detect, tested_depths = tested_depths, tested_probs = tested_probs, final_results = final_results))
}

```



```{r}
#| label: setup-functions

find_total_cost <- function(seq_depth, cost_of_seq, cost_of_proc, cost_of_sample, num_samples) {
  return(seq_depth * (cost_of_seq + cost_of_proc) + cost_of_sample * num_samples)
}

```



```{r}
#| label: load-data

# load in data
data <- read_csv("../data/new-hiv-data.csv", show_col_types = FALSE)

our_data <- data %>% 
  mutate(hiv_ra = hiv_reads / read_depth)

our_data

# Find the mew factor by taking the average of hiv ra over all HIV people.
#mew <- our_data %>% summarize(mean = mean(hiv_ra)) %>% pull(mean)

mew <- our_data %>% 
  summarize(geo_mean = exp(mean(log(hiv_ra), na.rm = TRUE))) %>% 
  pull(geo_mean)


```
This study has 3 subjects, who provide us with mew of 5.59e-5, calculated via geometric mean.

Using this, along with the assumption of a population of 300 million people, 100 reads as our detection threshold, a growth rate of 0.01336 which we derived from existing Nature study (source: https://www.nature.com/articles/nature19827), initial infection population of 100. We run simulations, where we vary the pool size, and cumulative incidence.


 
```{r}
#| label: run-analysis-temp

# Run analysis to get detection distribution
#cat("Analyzing detection distribution...\n")
#results <- analyze_detection_distribution(r, I1, P, D, theta, mew, shedding_weeks, N, n_sims)
#
### Display summary statistics
#cat("\nDetection Distribution Summary:\n")
#cat(sprintf("Detection rate: %.2f%%\n", results$detection_rate * 100))
#if (results$detection_rate > 0) {
#  cat(sprintf("Mean cumulative incidence at detection: %.2f%%\n", results$summary_stats$mean_incidence))
#  cat(sprintf("Median cumulative incidence at detection: %.2f%%\n", results$summary_stats$median_incidence))
#  cat(sprintf("IQR: %.2f%% - %.2f%%\n", results$summary_stats$q25_incidence, results$summary_stats$q75_incidence))
#  cat(sprintf("Range: %.2f%% - %.2f%%\n", results$summary_stats$min_incidence, results$summary_stats$max_incidence))
#}
#
### Calculate detection probabilities at various cumulative incidence thresholds
#thresholds <- c(1.0, 5.0, 10.0)  # % cumulative incidence
#cat("\n\nDetection Probabilities at Different Cumulative Incidence Levels:\n")
#for (threshold in thresholds) {
#  p_detect <- calc_detection_probability(results, threshold)
#  cat(sprintf("P(detect | %.1f%% cumulative incidence) = %.2f%%\n", threshold, p_detect * 100))
#}
```

```{r}
#| label: find-seq-depth

## Model parameters
# CONSTANTS
N <- 300e6  # US population
theta <- 100  # detection threshold (reads)
shedding_weeks <- 12
r <- 0.0136      # weekly growth based on HIV
I1 <- 100     # 100 initial infections


# User defined parameters
P <- 1000  # pool size
D <- 1e7   # sequencing depth to test (10 million reads)
n_sims <- 1000  # number of simulations

# Find optimal sequencing depths for different target incidences
target_incidences <- c(1, 5, 10)  # Different cumulative incidence targets
pool_sizes <- c(16, 64, 128, 256, 512, 1024, 2048)  # Different cumulative incidence targets
optimal_results <- list()


# Parallel
workers <- max(1, parallel::detectCores() - 1)
plan(multisession, workers = workers)

param_grid <- expand.grid(
  target_incidence = target_incidences,
  pool_size       = pool_sizes,
  stringsAsFactors = FALSE
)

#optimal_results <- future_apply(
#  X   = param_grid,          # each row is one job
#  MARGIN = 1,
#  FUN = function(row) {
#    find_optimal_depth(
#      target_incidence = row["target_incidence"],
#      target_prob      = 0.95,
#      r = r, I1 = I1, P = row["pool_size"],
#      theta = theta, mew = mew,
#      shedding_weeks = shedding_weeks,
#      N = N, n_sims = 500
#    )
#  },
#  future.seed = TRUE         # reproducible RNG streams
#)

# After the future_apply() call finishes
#results_dir <- "../results"        # adjust as needed
#if (!dir.exists(results_dir)) dir.create(results_dir)
#
#saveRDS(optimal_results,
#       file = file.path(results_dir, "optimal_results.rds"))

optimal_results <- readRDS("../results/optimal_results.rds")





#for (i in target_incidences) {
#for (j in pool_sizes) {
#  cat(sprintf("\n--- Target: %.1f%% cumulative incidence ---\n", target_incidences[i]))
#  optimal_results[[i]] <- find_optimal_depth(
#    target_incidence = target_incidences[i],
#    target_prob = 0.95,
#    r = r, I1 = I1, P = j, theta = theta, 
#    mew = mew, shedding_weeks = shedding_weeks, 
#    N = N, n_sims = 500
#  )
#}
#}

optimal_summary <- param_grid %>%         # nrow = length(optimal_results)
  mutate(
    optimal_depth = map_dbl(optimal_results, "optimal_depth"),
    achieved_prob = map_dbl(optimal_results, "final_prob"),
    q25_incidence = map_dbl(optimal_results,
                            c("final_results", "summary_stats", "q25_incidence")),
    q75_incidence = map_dbl(optimal_results,
                            c("final_results", "summary_stats", "q75_incidence")),
    annual_reads  = optimal_depth * 52
  )

optimal_summary <- as_tibble(optimal_summary)


ggplot(optimal_summary,
       aes(x = target_incidence,
           y = optimal_depth,
           colour = factor(pool_size),
           group  = pool_size)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
  scale_colour_brewer(palette = "Accent", name = "Pool size") +
  labs(title = "Weekly sequencing depth",
       y = "Reads per week",
       x = "Cumulative incidence",
       colour = "Pool size") +
  theme_light(base_size = 13) +
  theme(
    plot.title      = element_text(hjust = .5, face = "bold"),
    plot.subtitle   = element_text(hjust = .5),
    legend.position = "top"
  )


```

```{r}
#| label: cost-analysis

## Cost analysis for a specific depth
seq_c <- 2500 / 1e9
proc_c <- 2500 / 1e9
sample_c <- 100

# 1. add a new column with the cost for each depth
optimal_summary <- optimal_summary %>% 
  mutate(total_cost_per_year = find_total_cost(
           seq_depth      = optimal_depth,
           cost_of_seq    = seq_c,
           cost_of_proc   = proc_c,
           cost_of_sample = sample_c,
           num_samples    = pool_size
         )*52)

ggplot(optimal_summary,
       aes(x = pool_size,
           y = total_cost_per_year,
           fill = factor(target_incidence))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
  scale_x_log10() +
  scale_colour_brewer(palette = "Accent", name = "Cumulative incidence") +
  labs() +
  theme_light(base_size = 13) +
  theme(
    plot.title      = element_text(hjust = .5, face = "bold"),
    plot.subtitle   = element_text(hjust = .5),
    legend.position = "top"
  )

#optimal_summary %>% select(target_incidence, pool_size, total_cost_per_year, q25_incidence,q75_incidence) %>%
#  filter(pool_size == 1024)

```

*Note: Need to increase upper bound of optimal depth search as we hit 1billion reads for 1% cumulative incidence, which is the upper limit.*


# @TODO 

- Write explanation
- Make more plots
- Make better plot
- Confirm cost model stuff with Lenni
