---
title: "Blood Cost modeling"
subtitle: "A brief introduction"
date: "2025-05-16"
author:
  - name: "Harmon Bhasin"
    affiliation: "SecureBio"
format:
  html:
    toc: true
    code-fold: true
    toc-depth: 2
    number-sections: true
    theme: cosmo
  pdf:
    toc: true
---

# Motivation

*I'd like to thank Dan and Simon for our conversations about cost modeling. This work heavily draws from [Simon's blog post on pooled nasal swabs](https://naobservatory.org/blog/investigating-the-sensitivity-of-pooled-swab-sampling-for-pathogen-early-detection/).*

We'd like to make a cost-estimate to run a blood biosurveillance system in the US for a year at a sequencing depth capable of detecting a novel-HIV-like pathogen at X% cumulative incidence.

For this model, my goal was to make the most simple possible solution.

# Background

Before modeling, let's consider the typical process of sequencing a virus within a population till we detect it:

- There is the general population, and the virus of interest is circulating in that population with some incidence
- We sample people from this population to get our pool
- For each person in the pool, there is some probability that they're infected with our virus, and them being infected with the virus will potentially result in some portion of reads from their blood being from our virus of interest
- We sequence this pool (or multiple pools) until we hit the number of reads that we determine to be sufficient for detection of the aforementioned virus

Given this framework, we can begin to make some modeling decisions and assumptions (we may later decide to relax some of these). First, we assume some fixed weekly incidence of our HIV-like virus. Using a weekly incidence allows us to estimate the cost of sequencing required in a week, which then allows us to get the annual number by multiplying the total cost per week by 52. A week is also a handy number as we'd probably want to be sampling around once a week. Using this fixed weekly incidence, combined with the [doubling rate per day for the HIV-like virus](https://bionumbers.hms.harvard.edu/bionumber.aspx?id=105728&ver=6) (note: this source is incorrect, i'll need to do more research to find a source that tells us the doubling rateof HIV, for now we'll assume that this is correct), we can also derive a fixed prevalence (for this, I use the [code](https://naobservatory.org/blog/investigating-the-sensitivity-of-pooled-swab-sampling-for-pathogen-early-detection/) provided in the pooled swab sampling post; to find this code, click on the first code block on the blog and find the `get_prevalence` function):

\begin{align*}
\text{growth factor} &= 2^{\frac{1}{\text{doubling period days}}} \\
\text{weekly growth} &= \text{growth factor}^{\text{days in week}} \\
\text{prevalence} &= \text{weekly incidence} \times \frac{\text{weekly growth}}{\text{weekly growth}-1}
\end{align*}

Second, we will estimate a ratio that will allow us to convert an HIV-infected person to the relative abundance HIV-like virus reads that we'd find from sequencing their blood. We will call this the *i2ra* ratio. To estimate this value, we will use existing metagenomic sequencing data from people with HIV.

We'd prefer to find studies where 1) confirmed HIV positive people, 2) individual samples, and 3) they provide the number of HIV reads, along with the total sequencing depth. I prefer the studies where they provide these directly in a table, as it's the fastest to process, and we don't have to justify the use of computational methods (these would be necessary if we tried to derive this using the raw sequencing data). In the case that we analyze multiple datasets (which is probably likely), we would want to account for the fact that the datasets use different methods and correct for that (it would be good to factor in large error bars), or analyze them separately, similar to how [p2ra does it](https://www.medrxiv.org/content/10.1101/2023.12.22.23300450v4).

Third, we assume that we'll be running sequencing on pooled samples. This will be cheaper than individual samples, and in some cases, we may not be able to get the samples individually.

# Methods

We estimate our annual cost by breaking the weekly cost into two parts, 1) sequencing based costs, and 2) per-sample collection cost, then summing them over all the weeks in a year:

\begin{align*}
\text{Total cost per week} &= \\
& \underbrace{\text{sequencing depth} \times (\text{cost of sequencing} + \text{cost of processing})}_{\text{Sequencing based costs}} + \\
&\underbrace{\text{cost of single sample} \times \text{number of samples}}_{\text{Per-sample collection cost}} \\
\text{Total cost per year} &= \text{Total cost per week} \times 52
\end{align*}

where:

- sequencing depth: the number of reads required to detect our HIV-like pathogen
- cost of sequencing: the cost of sequencing a pool of samples using an arbitrary sequencing platform
- cost of processing: the cost of sample preparation and library preparation
- cost of single sample: the cost of purchasing and shipping a single sample
- number of samples: the number of samples we get; this will be our pool size

Of these variables, the main one we have to model is sequencing depth. The rest can be collected from the literature or other sources. I first introduce some notations:

- $ps$: pool size (user-defined)
- $i$: number of infected people
- $i2ra$: infected person to relative abundance ratio
- $n$: sequencing depth
- $d$: detection threshold (user-defined)
- $pr$: prevalence
- $ra$: relative abundance

Given this notation, we can find the sequencing depth to detect our HIV-like pathogen using the following equation:

$$ n = \frac{d}{(\frac{i}{ps}) \times i2ra}$$

Since there are some non-deterministic factors that will affect the sequencing depth, we'll run multiple simulations to calculate our sequencing depth, which will then provide us with a distribution of sequencing depths. We can then use the median sequencing depth in the "Total cost per week" equation above (*note: one thing to add is confidence intervals to this estimate*).

Below I will discuss how to calculate the *i2ra* ratio using pre-existing untargeted metagenomic sequencing data from people with HIV, then show how to calculate the sequencing depth using the *i2ra* ratio.

## Estimating the infected person to relative abundance ratio (*i2ra*)

To estimate the *i2ra* rato for our HIV-like pathogen, we can take an existing untargeted metagenomic sequencing study that was done on people with confirmed HIV and find the HIV relative abundance by diving the HIV read count by the total read count for those with HIV(before cleaning of course). We can then find the average HIV relative abundance over those individuals with HIV to get our *i2ra* ratio:

$$ i2ra = \frac{1}{M} \sum_{m \in \text{HIV individuals}}^M \frac{m_{\text{HIV read count}}}{m_{\text{Total read count}}} $$

## Estimating sequencing depth (*n*)

Estimating sequence depth requires 1) estimating the number of infected people in the pool, 2) calculating the relative abundance of HIV in those pools, and 3) using the read threshold with the relative abundance to get the sequencing depth.

### Estimating the number of infected people in the pool (*i*)

Instead of simply multiplying prevalence by pool size, which can misleadingly suggest zero infections when prevalence is low (e.g., with 1% prevalence in a pool of 50, the average is 0.5, not 0, however we cannot have 0.5 people infected), we use the Poisson distribution as it provides a more realistic model by calculating the probabilities of finding different numbers of infected individuals in the pool as rare, independent events. This approach is particularly useful when dealing with large pools and low infection rates. Hence the number of infected people in a pool can be represented by:

$$ i \sim Poisson(\lambda=pr \times ps) $$

### Estimating the relative abundance of our HIV-like pathogen in the pool (*ra*)

We calculate the relative abundance of the pool by applying the *i2ra* factor to the fraction of the pool that is infected:

$$ ra = \frac{i}{ps} \times i2ra $$

*Note: We may want to take a more rigorous approach here by using a distribution to model the relative abundance of the pathogen. More details on this can be found in the section labeled improvements.*

### Estimate the required sequencing depth using the read threshold (*d*)

To figure out sequencing depth, we can divide the read threshold by the relative abundance of the pathogen in the pool:

$$ n = \frac{d}{ ra } $$

# Results

To test the validity of the model, and get some initial results, I decided to analyze [one study](https://pmc.ncbi.nlm.nih.gov/articles/PMC3807392/). This study looked at how AIDS alters the plasma virome (*note: for the final dataset(s) we're going to focus on people earlier on in HIV infection status; see the section labeled "Consider the timing of infection in the *i2ra* estimate" under the "Improvements" header to understand why*), and it is a perfect candidate according to the reasons I laid out above, namely that it fits all three criteron, and this information can be found in Table 1 of the paper. For this analysis, I only analyzed the samples of all people not on ART, since it appears to [change the rate of HIV nucleic acid found in blood](https://pmc.ncbi.nlm.nih.gov/articles/PMC4392820/), which may make the HIV read count lower than it actually is. Additionally, I use the standardized viral count for HIV that they provide in Table 1, we will probably want the unstandardized viral count if we do decide to move forward with this dataset.

```{r}
#| label: load-packages
#| output: false

library(tidyverse)

```

```{r}
#| label: setup-functions

# Define function to calculate read depth, and store some summary statistics
# 1. Use poisson distribution to estimate the number of infected people
# 2. Calculate ra based on infected people in pool
# 3. Calculate total number of reads for detection

find_seq_depth <- function(prevalence, i2ra_factor, pool_size, detection_threshold, simulations) {
  results <- data.frame(
    n_infected = numeric(simulations),
    expected_ra = numeric(simulations),
    required_depth = numeric(simulations)
  )
  for (i in 1:simulations) {
    n_infected <- rpois(1, pool_size * prevalence)
    if (n_infected == 0) {
      # Can't divide by 0
      expected_ra <- 1e-15
    } else {
      expected_ra <- (n_infected / pool_size) * i2ra_factor
    }
    required_depth <- detection_threshold / expected_ra
    results$n_infected[i] <- n_infected
    results$expected_ra[i] <- expected_ra
    results$required_depth[i] <- required_depth
  }
  summary_stats <- list(
    mean_infected = mean(results$n_infected),
    median_infected = median(results$n_infected),
    zero_pools_pct = 100 * mean(results$n_infected == 0),
    mean_ra = mean(results$expected_ra[results$n_infected > 0]),
    median_ra = median(results$expected_ra[results$n_infected > 0]),
    mean_depth = mean(results$required_depth[is.finite(results$required_depth)]),
    median_depth = median(results$required_depth[is.finite(results$required_depth)]),
    depth_95pct = quantile(results$required_depth[is.finite(results$required_depth)], 0.95)
  )
  return(list(results = results, summary = summary_stats))
}

# Calculate total cost

find_total_cost <- function(seq_depth, cost_of_seq, cost_of_proc, cost_of_sample, num_samples) {
  return(seq_depth * (cost_of_seq + cost_of_proc) + cost_of_sample * num_samples)
}

```

Using Gemini, I copied the entries from Table 1 of the paper to a tsv file locally called `hiv-data.csv`. Should we continue with this study, I may provide more information on the sample processing, etc, that they used.

```{r}
#| label: load-data

# load in data
data <- read_csv("../data/hiv-data.csv")
# select columns of interest
our_data <- data %>% select("HIV", "HIV load (RNA copies/ml)", "Total no. of reads", "No. of unique reads", "ART")
our_data <- our_data %>% rowid_to_column("sample_id")
# subset to patience of interest and calculate ra of hiv
our_data <- our_data %>% filter(ART == "No") %>% mutate("hiv_ra" = HIV / `Total no. of reads`)

# Find the i2ra factor by taking the average of hiv ra over all HIV people.
i2ra <- our_data %>% summarize(mean = mean(hiv_ra)) %>% pull(mean)

```

Of the 35 subjects in this study, all of whom have HIV, I only analyzed the 14 that were not taking ART. Of those 14 people, 4 people had undetectable levels viral nucleic acid present in their blood. We get an i2ra factor of 7e-5.


Using a 1% weekly incidence with a doubling rate of [0.65](https://bionumbers.hms.harvard.edu/bionumber.aspx?id=105728&ver=6), we get a prevalence of ~1%. We can then set the read threshold of detection to be 100, try a pool size of 100, and run 1000 runs of the simulation  This gives us a required median sequencing depth of about 1.34 × 10^8 or ~13 million reads per week.

```{r}
#| label: find-seq-depth

static_prevalence <- FALSE
if (static_prevalence) {
  prevalence <- 0.01
} else {
  # Inherent to pathogen
  doubling_period_days <- 0.65
  # We set this
  weekly_incidence <- 0.01
  get_prevalence <- function() {
    days_in_week <- 7
    growth_factor <- 2 ^ (1 / doubling_period_days)
    weekly_growth <- growth_factor ^ days_in_week
    prevalence <- weekly_incidence * (weekly_growth) / (weekly_growth - 1)
    return(prevalence)
  }
  prevalence <- get_prevalence()
}
prevalence

pool_size <- 100
detection_threshold <- 100
number_of_simulations <- 1000

# Get sequencing depth
seq_depth_results <- find_seq_depth(
  prevalence = prevalence,
  i2ra_factor = i2ra,
  pool_size = pool_size,
  detection_threshold = 100,
  simulations = number_of_simulations
)

seq_c <- 2500 / 1e9
proc_c <- 2500 / 1e9
seq_depth <- seq_depth_results$summary$median_depth
seq_depth

weekly_cost <- find_total_cost(
  seq_depth = seq_depth,
  cost_of_seq = seq_c,
  cost_of_proc = proc_c,
  cost_of_sample = 100,
  num_samples = pool_size
)

yearly_cost <- weekly_cost * 52
yearly_cost

```

To calculate the total cost, we use the numbers from the [p2ra paper](https://www.medrxiv.org/content/10.1101/2023.12.22.23300450v4) for the sequencing cost and processing costs, which comes out to be about $2500 each for 1 billion read pairs. For the blood sample costs, I use the higher end of [our estimate](https://docs.google.com/document/d/1rF5fCTDDqa0lZ8Fms9forA-pHMsB2fCcr3dn9ynrW1U/edit?tab=t.0), which puts us at about $100 per sample. Plugging this into our equation with the sequencing depth, we get a cost of about 55 thousand dollars per year, nice.

Hence, given a read threshold of 100, a pool size of 100, and a weekly incidence of 1%, it would cost about 55k per year to detect an HIV-like pathogen. (*note: we'll use confidence intervals for the final calculation so that this is a range*; silly question for Dan or Simon: how would I convert this weekly incidence number to the cumulative incidence?)

# Improvements

## Consider the timing of infection in the *i2ra* estimate

From Lenni:

I think this mostly makes sense but one nuance to consider here is timing of infection. If we want to detect an HIV-like pathogen early, most people will be donating early in the course of infection. HIV has very different shedding dynamics depending on time. Averaging over these states seems mostly reasonable but ideally we'd find people early in the course of infection, asymptomatic, and not taking anti-virals



## Simulating sampling multiple times a week

Right now we simulate sampling once a week. Lenni suggested that we may want to sample more than this (e.g. 2x-3x per week), it would be to figure out what would need to change to allow for this.

## Make some plots

This would be nice to have.

## Adding more studies

We will most likely want to get more data for estimating the i2ra factor.

## Plot a sweep of different weekly incidences and pool sizes

It would be nice to see how pool size and weekly incidence affect the cost. We could make a plot similar to what is done in the p2ra paper.

## Model prevalence as a function of time

If we wanted to model an outbreak, it would be better to assume a non-constant prevalence.

## Use a longer time horizon for incidence

Right now we use weekly incidence as this is what both the nasal swab and wastewater paper do. We will want to decide whether this is good for us.

## Estimate relative abundance of pool using a distribution

You may have noticed that I look at the median sequencing depth instead of the mean sequencing depth from the simulation. This is because we often have 0 infected people in the pool, however this leads to issues as, we cannot divide by 0. We may decide to model the relative abundance of the pool as a distribution similar to Simon's nasal swab work. 

## Double check all numbers

Double check all the numbers, and make sure they make sense.

# Appendix

## Sanity checking modeling

Just for sanity check, we can set the seq_depth to be 8e5, similar to norovirus example in P2ra, paper and we find that we get similar ball park.

Rothman relative abundance for norovirus GII seems to be about 1e-4.

> The number of reads required varies substantially between different viruses and studies. For untargeted sequencing, we find that detection at 1% cumulative incidence, with a read threshold of 100 reads, requires 8·7 × 10^5 ... sequencing reads for norovirus GII Rothman.
> At 1% TCI, the predicted annual cost to detect a GII norovirus-like pathogen under Rothman ... was $52,200.

```{r}
#| label: sanity-check

seq_c <- 2500 / 1e9
proc_c <- 2500 / 1e9
seq_depth <- 8e5

weekly_cost <- find_total_cost(
  seq_depth = seq_depth,
  cost_of_seq = seq_c,
  cost_of_proc = proc_c,
  cost_of_sample = 10,
  num_samples = pool_size
)

yearly_cost <- weekly_cost * 52
```

This gives us a similar cost, of about 52k per year, so we know that our cost modeling is not too far off.
