---
title: "Blood Cost modeling"
subtitle: "A brief introduction"
date: "2025-05-16"
author:
  - name: "Harmon Bhasin"
    affiliation: "SecureBio"
format:
  html:
    toc: true
    code-fold: true
    toc-depth: 2
    number-sections: true
    theme: cosmo
  pdf:
    toc: true
---

# Motivation

*I'd like to thank Dan and Simon for our conversations about cost modeling. This work heavily draws from [Simon's blog post on pooled nasal swabs](https://naobservatory.org/blog/investigating-the-sensitivity-of-pooled-swab-sampling-for-pathogen-early-detection/).*

We'd like to estimate the cost of running a blood biosurveillance system in the US for a year that is capable of detecting a HIV like-pathogen at X% weekly incidence. Specifically, we want to know the sensitivity of blood to detect such a pathogen. For this model, my goal was to make the most simple solution, as this will be used in a perspective piece and will only take up one section of the main text (derivations would go in the appendix).

# Background

**We want to know the annual sequencing depth required to detect a pathogen at X% cumulative incidence with a $r$ weekly growth rate.**

## Key Model Assumptions

1. **Exponential Growth**: Infections grow deterministically exponentially to reach cumulative incidence $X$% with growth rate $r$.
2. **Fixed Shedding**: Each infected person sheds for exactly 12 weeks at constant rate
3. **Random Sampling**: Each week, pool represents random sample from total population. Probability of getting sampled is independent of your probability of getting sampled.
4. **Constant mew**: All infected individuals contribute same relative abundance $mew$
5. **Poisson Reads**: Read counts follow Poisson distribution around expected value
6. **Independent Weeks**: Sampling is independent across weeks (with replacement)

## Modeling

### 1. Exponential Growth Model

The number of new infections in week $t$ follows exponential growth:

$$I(t) = I_1 \cdot e^{r(t-1)}$$

where $I_1$ is the initial number of infections in week 1, and $r$ is the weekly growth rate.

**Constraint**: The total cumulative infections must equal the target:

$$\sum_{t=1}^{T} I(t) = \frac{X}{100} \cdot N$$

**Solving for parameters**: Given the exponential sum formula:

$$\sum_{t=1}^{T} I_1 \cdot e^{r(t-1)} = I_1 \cdot \frac{e^{rT} - 1}{e^r - 1} = \frac{X \cdot N}{100}$$

Given the growth rate $r$ cumulative incidence $X$ and initial infection number $I_1$ , we can solve for the time period $T$ in weeks, that it would take for the pathogen to hit $X$ cumulative incidence.
### 2. Shedding Population

The number of people shedding in week t should be those who were infected in the **previous** 12 weeks:

$$S(t) = \sum_{k=\max(1, t-11)}^{t} I(k)$$

This accounts for the 12-week shedding period after infection.

### 3. Sampling Process

The key insight is that **we're sampling without replacement from a finite population with two types of people** (shedding vs. non-shedding). 

Each week, we randomly sample $P$ people from the total population $N$. The number of shedding individuals captured in our sample follows a hypergeometric distribution:

$$\text{Sampled shedders} \sim \text{Hypergeometric}(N, S(t), P)$$

The probability mass function is:

$$P(\text{sampled shedders} = k) = \frac{\binom{S(t)}{k} \binom{N-S(t)}{P-k}}{\binom{N}{P}}$$

### 4. Read Detection

The expected number of pathogen reads detected in week $t$ is:

$$E[R(t)] = \text{(sampled shedders)} / P \times i2ra \times D$$

The actual reads follow a Poisson distribution around this expectation:

$$R(t) \sim \text{Poisson}(\text{(sampled shedders)} / P \times i2ra \times D)$$

where D is our sequencing depth.
### 5. Detection Criterion

Detection occurs when cumulative reads over all weeks exceed the threshold:

$$R_{cum} = \sum_{t=1}^{T} R(t) \geq \theta$$

### 6. Detection Probability

For a given set of parameters $(X, P, D)$, the detection probability is:

$$p_{detect}(X, P, D) = P(R_{cum} \geq \theta)$$

This is estimated through Monte Carlo simulation due to the complex stochastic dependencies.

## Objective Function

The goal is to find the minimum sequencing depth $D^*$ that achieves a target detection probability $p_{target}$ (typically 95%):

$$D^* = \min{D : p_{detect}(X, P, D) \geq p_{target}}$$
Once the optimal weekly depth $D*$ is found, annual sequencing demand is $D^*_{yr} = 52 \times D*$

**Second**, we estimate the expected relative abundance of the HIV like-pathogen for an infected person by using existing metagenomic sequencing data from people with HIV (NB: this should become "people early in their infection with HIV" once we change the data). We call this the *i2ra* ratio.

We look for studies where 1) people were confirmed to be HIV positive, 2) these individual's samples were sequenced separately, 3) they provide the number of HIV reads, along with the total sequencing depth, and 4) optionally, they are early in their infection of HIV. We prefer the studies where they provide this information in a table, as it's the fastest to process, and doesn't require us running their data through our pipeline (NB: we will need to highlight the differences between the studies methods if we decide to go with multiple). In the case that we analyze multiple datasets (which is probably likely), we would want to account for the fact that the datasets use different methods and correct for that (it would be good to factor in large error bars), or analyze them separately, similar to how [p2ra does it](https://www.medrxiv.org/content/10.1101/2023.12.22.23300450v4).

# Cost modeling

We estimate our annual cost by breaking the weekly cost into two parts, 1) sequencing based costs, and 2) per-sample collection cost, then summing them over all the weeks in a year:

\begin{align*}
\text{Total cost per week} &= \\
& \underbrace{\text{sequencing depth} \times (\text{cost of sequencing} + \text{cost of processing})}_{\text{Sequencing based costs}} + \\
&\underbrace{\text{cost of single sample} \times \text{number of samples}}_{\text{Per-sample collection cost}} \\
\text{Total cost per year} &= \text{Total cost per week} \times 52
\end{align*}

where:

- sequencing depth: the number of reads required to detect our HIV-like pathogen
- cost of sequencing: the cost of sequencing a pool of samples using an arbitrary sequencing platform
- cost of processing: the cost of sample preparation and library preparation
- cost of single sample: the cost of purchasing and shipping a single sample
- number of samples: the number of samples we get; this will be our pool size (NB: this is an error that I need to fix; right now I assume number of samples is pool size)

# Results

To test the validity of the model, and get some initial results, I decided to analyze [one study](https://pmc.ncbi.nlm.nih.gov/articles/PMC3807392/). This study looked at how AIDS alters the plasma virome (*NB: for the final dataset(s) we're going to focus on people earlier on in HIV infection status; see the section labeled "Consider the timing of infection in the *i2ra* estimate" under the "Improvements" header to understand why*), and it is a perfect candidate according to the reasons I laid out above, namely that it fits all three criteron, and this information can be found in Table 1 of the paper. For this analysis, I only analyzed the samples of all people not on ART, since it appears to [change the rate of HIV nucleic acid found in blood](https://pmc.ncbi.nlm.nih.gov/articles/PMC4392820/), which may make the HIV read count lower than it actually is. Additionally, I use the standardized viral count for HIV that they provide in Table 1, we will probably want the unstandardized viral count if we do decide to move forward with this dataset.

```{r}
#| label: load-packages
#| output: false

library(tidyverse)
library(furrr)
library(dplyr)

```

```{r}
#| label: aux-functions

# Function to solve for T given r, I1, and X
solve_time_period <- function(r, I1, X_pct, N) {
  target_total <- (X_pct / 100) * N
  if (r == 0) {
    # Constant infections case
    return(target_total / I1)
  } else {
    # Exponential growth case
    # T = (1/r) * ln(1 + (X*N*(e^r - 1))/(100*I1))
    exp_r <- exp(r)
    ratio <- (X_pct * N * (exp_r - 1)) / (100 * I1)
    T <- (1/r) * log(1 + ratio)
    return(T)
  }
}

# Function to calculate infections at time t
calc_infections <- function(t, I1, r) {
  I1 * exp(r * (t - 1))
}

# Function to calculate shedding population at time t
calc_shedding <- function(t, I1, r, shedding_weeks) {
  start_week <- max(1, t - shedding_weeks + 1)
  end_week <- t
  weeks <- start_week:end_week
  infections <- sapply(weeks, function(w) calc_infections(w, I1, r))
  sum(infections)
}

```

```{r}
#| label: simulate-epidemic

# Function to simulate one epidemic scenario
simulate_epidemic <- function(r, I1, X_pct, P, D, n_sims = 1000) {
  # Solve for time period T
  T <- solve_time_period(r, I1, X_pct, N)
  T <- ceiling(T)  # Round up to whole weeks
  if (T <= 0) {
    warning("Invalid time period calculated")
    return(0)
  }
  # Pre-calculate shedding population for each week
  shedding_pop <- sapply(1:T, function(t) {
    calc_shedding(t, I1, r, shedding_weeks)
  })
  # Run simulations
  detections <- replicate(n_sims, {
    total_reads <- 0
    for (t in 1:T) {
      # Sample from population using hypergeometric
      S_t <- round(shedding_pop[t])
      S_t <- min(S_t, N)  # Cap at population size
      if (S_t > 0 && P < N) {
        # For large populations, approximate with binomial
        sampled_shedders <- rbinom(1, P, S_t / N)
        # Generate reads from Poisson distribution
        expected_reads <- sampled_shedders / P * i2ra * D
        weekly_reads <- rpois(1, expected_reads)
        total_reads <- total_reads + weekly_reads
      }
    }
    total_reads >= theta
  })
  mean(detections)
}

```

```{r}
#| label: calculate-depths

# function to find minimum sequencing depth
find_min_depth <- function(r, i1, x_pct, p, n, theta, i2ra, shedding_weeks, target_prob = 0.95) {
  # binary search for minimum depth
  d_low <- 1e6
  d_high <- 1e9
  tolerance <- 0.01
  while ((d_high - d_low) / d_low > tolerance) {
    d_mid <- (d_low + d_high) / 2
    prob <- simulate_epidemic(r, i1, x_pct, p, d_mid, n_sims = 500)
    if (prob >= target_prob) {
      d_high <- d_mid
    } else {
      d_low <- d_mid
    }
  }
  return(d_high)
}

# Function to validate parameters and show epidemic dynamics
validate_scenario <- function(r, I1, X_pct, N) {
  T <- solve_time_period(r, I1, X_pct, N)
  T_weeks <- ceiling(T)
  # Calculate total infections to verify
  total_infections <- I1 * (exp(r * T_weeks) - 1) / (exp(r) - 1)
  target_infections <- (X_pct / 100) * N
  cat("Scenario validation:\n")
  cat(sprintf("Growth rate (r): %.3f per week\n", r))
  cat(sprintf("Initial infections (I1): %.0f\n", I1))
  cat(sprintf("Target cumulative incidence: %.1f%%\n", X_pct))
  cat(sprintf("Calculated time period: %.1f weeks\n", T))
  cat(sprintf("Rounded time period: %d weeks\n", T_weeks))
  cat(sprintf("Target total infections: %.0f\n", target_infections))
  cat(sprintf("Actual total infections: %.0f\n", total_infections))
  cat(sprintf("Difference: %.1f%%\n", 100 * (total_infections - target_infections) / target_infections))
  return(T_weeks)
}

# Function to analyze single scenario
analyze_scenario <- function(r, I1, X_pct, N, theta, i2ra, shedding_weeks, target_prob = 0.95, pool_sizes = seq(1000, 5000, 500)) {
  # Validate scenario first
  T_weeks <- validate_scenario(r, I1, X_pct, N)
  # Setup parallel processing
  plan(multisession, workers = 4)
  # Analyze different pool sizes
  results <- tibble(P = pool_sizes) %>%
    mutate(
      min_depth = future_map_dbl(
        P,
        ~find_min_depth(r, I1, X_pct, .x, N, theta, i2ra, shedding_weeks, target_prob),
        .options = furrr_options(seed = 42)
      ),
      annual_depth = min_depth * 52,
    )
  return(results)
}

```



```{r}
#| label: setup-functions

find_total_cost <- function(seq_depth, cost_of_seq, cost_of_proc, cost_of_sample, num_samples) {
  return(seq_depth * (cost_of_seq + cost_of_proc) + cost_of_sample * num_samples)
}

```

To transfer the data from the paper (Table 1) to a tsv file, I copy pasted the text into Gemini and asked it to format the data as a tsv file. Should we continue with this study, I may provide more information on the sample processing, etc, that they used.


```{r}
#| label: load-data

# load in data
data <- read_csv("../data/new-hiv-data.csv")

our_data <- data %>% 
  mutate(hiv_ra = hiv_reads / read_depth)

ggplot(our_data, aes(x = hiv_ra)) +
  geom_histogram(binwidth = 0.00002, color = "black", fill = "skyblue") +
  labs(
    title = "Distribution of HIV Relative Abundance (hiv_ra)",
    x = "HIV Relative Abundance (hiv_ra)",
    y = "Count"
  ) +
  theme_minimal()


# Find the i2ra factor by taking the average of hiv ra over all HIV people.
i2ra <- our_data %>% summarize(mean = mean(hiv_ra)) %>% pull(mean)

```
Of the 35 subjects in this study, all of whom have HIV, I only analyzed the 14 that were not taking ART. Of those 14 people, 4 people had undetectable levels viral nucleic acid present in their blood. Using all 14 people (including those with undetectable levels of HIV viral nucleic acid), and taking the arithmetic mean (NB: will use geometric mean going forward), we get an i2ra factor of 7e-5.

Using a 1% weekly incidence with a doubling rate of 362 days (see Appendix: Doubling rate per day for HIV-like pathogen), we get a prevalence of ~75% (nb: this is quite high, for covid-19 it was 1.2% at a 3 day doubling rate; this is a bit unintuitive for me). We can then set the read threshold of detection to be 100, try a pool size of 100, and run 1000 runs of the simulation. This gives us a required median sequencing depth of about 1.79 × 10^6 or ~1.79 million reads per week.

```{r}
#| label: find-seq-depth

## Model parameters
# CONSTANTS
N <- 300e6  # US population
theta <- 100  # detection threshold (reads)
shedding_weeks <- 12
target_prob <- 0.95
r <- 0.0136      # weekly growth based on HIV
I1 <- 100     # 1 initial infections gave a period of 782 weeks at 1% cumulative incidence, so we'll start with 100 gives a period of 434 weeks at 1$ cumulative incidence


# User defined
X_pct <- 1.0  # 1% cumulative incidence
pool_sizes <-  c(1000, 5000, 10000) # pool sizes to simulate

# Example usage function
run_example <- function(r, I1, X_pct, N, theta, i2ra, shedding_weeks, target_prob = 0.95, pool_sizes = seq(1000, 5000, 500)) {
  # Example parameters
  cat("Running example analysis...\n")
  results <- analyze_scenario(r, I1, X_pct, N, theta, i2ra, shedding_weeks, target_prob, pool_sizes)
  cat("\nResults:\n")
  print(results)
  return(results)
}

run_example(r, I1, X_pct, N, theta, i2ra, shedding_weeks, target_prob, pool_sizes)

seq_c <- 2500 / 1e9
proc_c <- 2500 / 1e9
seq_depth <- seq_depth_results$summary$median_depth
seq_depth

weekly_cost <- find_total_cost(
  seq_depth = seq_depth,
  cost_of_seq = seq_c,
  cost_of_proc = proc_c,
  cost_of_sample = 100,
  num_samples = pool_size
)

yearly_cost <- weekly_cost * 52
yearly_cost
```

To calculate the total cost, we use the numbers from the [p2ra paper](https://www.medrxiv.org/content/10.1101/2023.12.22.23300450v4) for the sequencing cost and processing costs, which comes out to be about $2500 each for 1 billion read pairs. For the blood sample costs, I use the higher end of [our estimate](https://docs.google.com/document/d/1rF5fCTDDqa0lZ8Fms9forA-pHMsB2fCcr3dn9ynrW1U/edit?tab=t.0), which puts us at about $100 per sample. Plugging this into our equation with the sequencing depth, we get a cost of about $520K per year.

Hence, given a read threshold of 100, a pool size of 100, and a weekly incidence of 1%, it would cost about 520k per year to detect an HIV-like pathogen.

# Improvements

## Consider the timing of infection in the *i2ra* estimate

From Lenni:

I think this mostly makes sense but one nuance to consider here is timing of infection. If we want to detect an HIV-like pathogen early, most people will be donating early in the course of infection. HIV has very different shedding dynamics depending on time. Averaging over these states seems mostly reasonable but ideally we'd find people early in the course of infection, asymptomatic, and not taking anti-virals

## Simulating sampling multiple times a week

Right now we simulate sampling once a week. Lenni suggested that we may want to sample more than this (e.g. 2x-3x per week), it would be to figure out what would need to change to allow for this.

## Make some plots

This would be nice to have.

## Adding more studies

We will most likely want to get more data for estimating the i2ra factor.

## Plot a sweep of different weekly incidences and pool sizes

It would be nice to see how pool size and weekly incidence affect the cost. We could make a plot similar to what is done in the p2ra paper.

## Model prevalence as a function of time

If we wanted to model an outbreak, it would be better to assume a non-constant prevalence.

## Use a longer time horizon for incidence

Right now we use weekly incidence as this is what both the nasal swab and wastewater paper do. We will want to decide whether this is good for us.

## Estimate relative abundance of pool using a distribution

You may have noticed that I look at the median sequencing depth instead of the mean sequencing depth from the simulation. This is because we often have 0 infected people in the pool, however this leads to issues as, we cannot divide by 0. We may decide to model the relative abundance of the pool as a distribution similar to Simon's nasal swab work. 

## Double check all numbers

Double check all the numbers, and make sure they make sense.

# Appendix

## Sanity checking modeling

Just for a sanity check, we can set the seq_depth to be 8e5. This is similar to norovirus example in P2RA paper.

Rothman relative abundance for norovirus GII seems to be about 1e-4.

> The number of reads required varies substantially between different viruses and studies. For untargeted sequencing, we find that detection at 1% cumulative incidence, with a read threshold of 100 reads, requires 8·7 × 10^5 ... sequencing reads for norovirus GII Rothman.
> At 1% TCI, the predicted annual cost to detect a GII norovirus-like pathogen under Rothman ... was $52,200.

```{r}
#| label: sanity-check

seq_c <- 2500 / 1e9
proc_c <- 2500 / 1e9
seq_depth <- 8e5

weekly_cost <- find_total_cost(
  seq_depth = seq_depth,
  cost_of_seq = seq_c,
  cost_of_proc = proc_c,
  cost_of_sample = 10,
  num_samples = pool_size
)

yearly_cost <- weekly_cost * 52
yearly_cost
```

This gives us a similar cost, of about 52k per year, so we know that our cost modeling is not too far off.

## Doubling rate per day for HIV-like pathogen {#doubling-rate-hiv}

For the purpose of epidemiological modeling, a crucial parameter is the **doubling time**—the time it takes for the number of infected individuals to double. To establish a scientifically-grounded value for a new, HIV-like pathogen, we must look to the early phase of the HIV epidemic in the 1970s and 1980s, before widespread awareness and effective treatments altered its natural course.

The standard epidemiological formula for this calculation is:

$$T_d = \frac{\ln(2)}{r}$$

Where:

* `T_d` is the **doubling time** for the year.
* `r` is the **intrinsic growth rate** of the epidemic per year.
* `ln(2)` is the natural logarithm of 2, a constant approximately equal to 0.693.

Luckily, a 2016 study from *Nature*, titled ["1970s and 'Patient 0' HIV-1 genomes illuminate early HIV/AIDS history in North America"](https://www.nature.com/articles/nature19827), found a range for this. The researchers empirically calculated the doubling time (T_d) to be 0.86 -> 1.12 years.

We can take the arithmetic mean of this to get:

$$T_d = 0.99 \text{ years}$$

Converting this to days gives us our final, justifiable figure:

$$0.99 \text{ years} \times 365 \frac{\text{days}}{\text{year}} \approx \textbf{362 days}$$

Based on the most direct empirical evidence from the early HIV epidemic, the justifiable doubling time for a new, unchecked HIV-like pathogen is approximately **362 days**.

For context, this is fundamentally different from a rapidly spreading respiratory virus. Wild-type SARS-CoV-2, for example, had a doubling time of approximately **[3 days](https://naobservatory.org/blog/investigating-the-sensitivity-of-pooled-swab-sampling-for-pathogen-early-detection/)** at the beginning of the pandemic. This stark contrast underscores the slow, persistent transmission dynamic of an HIV-like pathogen compared to the explosive growth of an airborne disease.


NB: It looks like we're just undoing this calculation to get the growth rate to calculate prevalence, I should double check this.
