---
title: "Blood preprint"
subtitle: "A brief introduction"
date: "2025-05-16"
author:
  - name: "Harmon Bhasin"
    affiliation: "SecureBio"
format:
  html:
    toc: true               # table of contents
    toc-depth: 2
    number-sections: true
    theme: cerulean
#  pdf:
#    toc: true
#    number-sections: true
#    colorlinks: true
---

# Motivation

We'd like to make a cost-estimate to operate such a system in the US for a year at a sequencing depth capable of detecting a novel-HIV-like pathogen at X% cumulative incidence.

# Background

*A lot of this has been heavily inspired by [Simon's blog post](https://naobservatory.org/blog/investigating-the-sensitivity-of-pooled-swab-sampling-for-pathogen-early-detection/).*

To think about this, it would be helpful to think about what's going on here. The overall process we're interested in can be defined as follows:

- There is the general population
- We sample from this population, to get some subset of the population
- This subset population has some rate at which it's infected with our virus
- We form pools from this subset of the population, where some people will be infected and some will not
- We sequence this pool until we hit some number of reads that we determine to be sufficient for detection

The main thing we need to do is figure out how deep we need to sequence to detect our pathogen.

We will use the framework from the swab pooling paper, which estimates sequencing depth using a weekly metric.

We will make a few assumptions:
- we know what the prevelance is (they determine this in the swab post), and it stays the same over time; hence we're taking a point estimate

To figure out how deep we need to sequence, we need to figure out the ratio of which an infected person has relative abundance, we will call this the i2ra, which will be our equvialent of the p2ra factor.

To do this, we'd prefer to find studies where 1) HIV positive people, 2) individual samples, and 3) they provide the number of HIV reads, along with the total sequencing depth. I prefer the studies where they provide these directly in a table, as it's the fastest, and we don't have to justify the use of computational methods. Of course, since they use different methods, we would proabbly need to just say that there will be differences, but using this should be good enough.

The study that [I chose](https://naobservatory.org/blog/investigating-the-sensitivity-of-pooled-swab-sampling-for-pathogen-early-detection/) was perfect:
- Table 1 gives us practically all the information we need
- Note: I used the standardized viral count for HIV; we probably want to have unstandardized viral count, however for this exploratory work this should be enough
- There were other studies that I looked at that we may consider in more depth, I just watned to get an intial study.

# Methods

## Estimating i2ra

To estimate the i2ra, I take our study, and filter the samples to all people not on ART, since it appears to decrease the amount of nucleic acid of HIV, which may make the HIV read count lower than it actually is. For these samples, I find the HIV relative abundance by diving the HIV read count by total read count. Using the HIV relative abudance for each individual, we take the average. This gives us the i2ra factor.

$$ i2ra = \sum (HIV reads / total reads) $$

## Estimating sequencing depth

To estimate sequencing depth, we take an approach similar to that of the nasal swab work.

Steps to sequencing depth:
1. Estimate the number of infected people in the pool
2. Calculate the relative abundance of HIV in that pool
3. Use relative abundance along with read threshold to get sequencing depth.

We now discuss each of these in a little bit more detail, but the formula boils down to:

$$ sequencingdepth = \frac{detectionthreshold}{(Poisson(\lambda=poolsize*prevalence)/poolsize) * i2ra}$$

### Estimating the number of infected people in the pool

To get the estimated number of infect people in a pool, we first need to figure out how many people in that pool are infected. To get that, we might naively multiply the prevalnce by the pool, however, often the prevlance, maybe be too small to get any peole in teh pool (e.g. prevalnce of 1% but pool of 50 would result in 0 infected). One might just round to the nearest number, however you would only ever get 1. Hence, what we can do is use a distribution. Specifically, we can use a Possion distribution, which helps account for this. Hence, number of infected people can be represeneted as 

$$ n_{infected} \sim Poisson(\lambda=poolsize*prevalence) $$

### Calcualte the relative abundance

To calculate the relative abundance, we can simply take the i2ra factor, and multiply it by the fraction of infected people. THis gives us a point estimate.

$$ ra = (n_{infected}/poolsize) * i2ra $$

### Estimate sequencing depth

To figure out sequencing depth, we can figure out total detection divided by ra to get teh sequencing depth.

$$ sequencingdepth = detectionthreshold / ra $$

## Running simulation

Since we have a non-deterministic portion of our sequencing depth, we can run a simulation to get a distribution of sequencing depth. If we were to switch any of the other estiamtes to be distrbiutions, we would also benefit form this as well. We run simulations so that we can hopefully converge to what the true expected value is rather than just choosing a point estimate.

In this case all that would be changing from estaimte to tesimate is the amount of infected people in the pool.

Going forward then, we can use the mean sequencing depth as the sequencing depth we're interested in.

## Cost estimate

We develop a simple cost model:

Total cost per week = ( mean sequencing depth * (cost of sequencing + cost of processing)) + (cost of single sample * pool_size)


# Analysis


```{r}
#| label: load-packages
#| output: false

library(tidyverse)
```

```{r}
#| label: load-data

# load in data
data <- read_csv("../data/hiv-data.csv")
our_data <- data %>% select("HIV", "HIV load (RNA copies/ml)", "Total no. of reads", "No. of unique reads", "ART")
our_data <- our_data %>% filter(ART == "No") %>% mutate("hiv_ra" = HIV / `Total no. of reads`)
our_data <- our_data %>% rowid_to_column("sample_id")

# Find the p2ra factor by taking the average of hiv ra over all HIV people.
b <- our_data %>% summarize(mean=mean(hiv_ra)) %>% pull(mean)
```

```{r}
#| label: define-functions

find_seq_depth <- function(prevalence, p2ra_factor, pool_size, detection_threshold = 10, simulations = 1000) {
  results <- data.frame(
    n_infected = numeric(simulations),
    expected_ra = numeric(simulations),
    required_depth = numeric(simulations)
  )
  for (i in 1:simulations) {
    n_infected <- rpois(1, pool_size * prevalence)
    if (n_infected == 0) {
      expected_ra <- 1e-15
    } else {
      expected_ra <- (n_infected/pool_size) * p2ra_factor
    }
    required_depth <- detection_threshold / expected_ra
    results$n_infected[i] <- n_infected
    results$expected_ra[i] <- expected_ra
    results$required_depth[i] <- required_depth
  }
  summary_stats <- list(
    mean_infected = mean(results$n_infected),
    median_infected = median(results$n_infected),
    zero_pools_pct = 100 * mean(results$n_infected == 0),
    mean_ra = mean(results$expected_ra[results$n_infected > 0]),
    median_ra = median(results$expected_ra[results$n_infected > 0]),
    mean_depth = mean(results$required_depth[is.finite(results$required_depth)]),
    median_depth = median(results$required_depth[is.finite(results$required_depth)]),
    depth_95pct = quantile(results$required_depth[is.finite(results$required_depth)], 0.95)
  )
  return(list(results = results, summary = summary_stats))
}

find_total_cost <- function(seq_depth, cost_of_seq, cost_of_proc, cost_of_sample, num_samples){
  return (seq_depth * (cost_of_seq + cost_of_proc) + cost_of_sample * num_samples)
}
```


```{r}
#| label: find-seq-depth

# Choose some parameters
prevalence <- 0.01
pool_size <- 100
detection_threshold <- 100
number_of_simulations <- 1000

# Get sequencing depth
seq_depth_results <- find_seq_depth(
  prevalence = prevalence,
  p2ra_factor = b,
  pool_size = pool_size,
  detection_threshold = 100,
  simulations = number_of_simulations
)

seq_c <- 2500 / 1e9
proc_c <- 2500 / 1e9
seq_depth <- seq_depth_results$summary$mean_depth

find_total_cost(
  seq_depth = seq_depth,
  cost_of_seq = seq_c,
  cost_of_proc = proc_c,
  cost_of_sample = 10,
  num_samples = pool_size
) * 52

# Just for sanity check, we can set the seq_depth to be 8e5, similar to norovirus example in P2ra, paper and we find that we get similar ball park.
#seq_depth <- 8e5

```



# Improvements

## Estimate prevalence not using point

Instead of having a set prevlance, we could use a prevalcne that grows with some time, as they do with the nasal swab work.


## Estimate relative abundance of pool using a distribution

Instead of taking an average, we could take a more displined approach to modeling the realtive abudnace fro a pool. In the nasal swab work, they fit a log normal distribution to do this.

Right now, we're hurt from infected people who have no viral nucleic acid, we may instead want to have a distribution for this.

We're assuming that the HIV study statistics is represetnative of the whole sample, however this is probably not true.


## Add more studies

We will most likely want to use more studies as this is just one study.


## Use the unstandardized viral count

To some extent, using the standaardized viral counts is going to hurt us.

