---
# Basic document metadata
title: "Blood preprint"
subtitle: "A brief introduction"
author:
  - name: "Harmon Bhasin"
    affiliation: "SecureBio"
format:
  html:
    toc: true               # table of contents
    toc-depth: 2
    number-sections: true
    theme: cerulean
#  pdf:
#    toc: true
#    number-sections: true
#    colorlinks: true

---

# Motivation

We’d like to make a cost-estimate model that would return the cost of detection for a system in the US for a year to detect a novel-HIV-like pathogen at X% cumulative incidence.


# Background


THigns to think about:
- is it an outbreak or is there some set prevelance (this is the difference between learning a point estimate vs takign it over time):
    - if outbreak then you do things based on some time steps 
    - if not then you chilling; you can just assume some thing 

The way we'd like to think about the problem is as follows:


- you have a population
- you sample that population
- the virus of interest has some prevelance
- sample a pool
- estimate the relative abudnace of the virus in the pool
- convert that to a number of sequencing reads
- plug that into the rest of the cost equation


# Data preparation

The preferred data we'd like are:
- study with HIV positive people
- individual samples
- number of HIV reads, along with the total sequencing depth


This provides the ideal data for our analysis as:
1. We don't have to analyze the data ourselves and provide an explanation for the analysis
2. Indiividal people, make the pooling calcualtion much easier
3. Having HIV positive people allow us to do the p2ra calculation



For this explaorty analysis I found one study (https://pmc.ncbi.nlm.nih.gov/articles/PMC3807392/):
- Table 1 gives us the information we need.
- I use the standardized viral count for HIV, and divdie it by the total number of reads
- i note that this is probably not right, but i just wanted some data to use.


# Analysis

First we can load in the table from the paper. I ported this table using Gemini, so we'll want to double check things, but this is a good start.

Given that we're intersted in an HIV-like pathogen, we proably don't want ART, as apparently it reduces the amount of viral copies found in the blood.

```{r}
#| label: load-data
library(tidyverse)

# load in data
data <- read_csv("../data/hiv-data.csv")
our_data <- data %>% select("HIV", "HIV load (RNA copies/ml)", "Total no. of reads", "No. of unique reads", "ART")
our_data <- our_data %>% filter(ART == "No") %>% mutate("hiv_ra" = HIV / `Total no. of reads`)
our_data <- our_data %>% rowid_to_column("sample_id")

# Find the p2ra factor by taking the average of hiv ra over all HIV people.
b <- our_data %>% summarize(mean=mean(hiv_ra)) %>% pull(mean)
# Assume a prevelance of 1%
prevalence <- 0.01
pool_size <- 100

```

Given the prevelance and p2ra factor, we can now run a simulation.

For the simulation, we can choose a pool of size that we'd like. We then can figure out what portion of people in that pool are infected. The naive version might just be round(pool_size * prevlanace), however, in the case where the prevalcne is small, this doens't wrok out too well as we always get 1. Hence, instead we can use a Poisson distribution because .... 

This then gives us the number of infected people in that pool. Given the number of infected people in that pool, we can then figure out what the expected relative abudnacne is by taking an average (n_infected/pool size) represetns portion of sample that has infection, then you can multiple that by the p2ra factor, this gives you the expected relative abundance, since you're going from prevlance in that pool to relative abdunacne. Then, what you have is the requried sequenceing depth.

What we can then do is run a simulation, when we're running the simulation, where we're getting the variation is from the poission distribution which will change each time, we can then take the average, and use that.
```{r}
#| label: prevelance

find_seq_depth <- function(prevalence, p2ra_factor, pool_size, detection_threshold = 10, simulations = 1000) {
  results <- data.frame(
    n_infected = numeric(simulations),
    expected_ra = numeric(simulations),
    required_depth = numeric(simulations)
  )
  for (i in 1:simulations) {
    n_infected <- rpois(1, pool_size * prevalence)
    if (n_infected == 0) {
      expected_ra <- 1e-15
    } else {
      expected_ra <- (n_infected/pool_size) * p2ra_factor
    }
    required_depth <- detection_threshold / expected_ra
    results$n_infected[i] <- n_infected
    results$expected_ra[i] <- expected_ra
    results$required_depth[i] <- required_depth
  }
  summary_stats <- list(
    mean_infected = mean(results$n_infected),
    median_infected = median(results$n_infected),
    zero_pools_pct = 100 * mean(results$n_infected == 0),
    mean_ra = mean(results$expected_ra[results$n_infected > 0]),
    median_ra = median(results$expected_ra[results$n_infected > 0]),
    mean_depth = mean(results$required_depth[is.finite(results$required_depth)]),
    median_depth = median(results$required_depth[is.finite(results$required_depth)]),
    depth_95pct = quantile(results$required_depth[is.finite(results$required_depth)], 0.95)
  )
  return(list(results = results, summary = summary_stats))
}

# Example usage
blood_analysis <- find_seq_depth(
  prevalence = prevalence,  
  p2ra_factor = b,   
  pool_size = pool_size,      
  detection_threshold = 100
)

```

Now that we have sequecning reads, we can plug this in to the rest of our equation 


We get the following 

Total cost per week = (sequencing depth * (cost of sequencing + cost of processing)) + (cost of sample * number of samples)

```{r}

seq_c <- 2500
proc_c <- 2500
#seq_depth <- blood_analysis$summary$mean_depth
seq_depth <- 8e5

find_total_cost <- function(seq_depth, cost_of_seq, cost_of_proc, cost_of_sample, num_samples){
  return (seq_depth * 1e-9 * (cost_of_seq + cost_of_proc) + cost_of_sample * num_samples)
}

find_total_cost(
  seq_depth = seq_depth,
  cost_of_seq = seq_c,
  cost_of_proc = proc_c,
  cost_of_sample = 10,
  num_samples = pool_size
) * 52

# Just for sanity check, we can set the seq_depth to be 8e5, similar to norovirus example in P2ra, paper and we find that we get similar ball park.

```





# Improvements 

- To get the RA of a pool of the HIV like virus, we take the average. A more displined approach might be sampling from a distribution
- Currently we assume some set prevelance. We may instead want to convert to an outbreak, this would make prevelance a funciton of t.
- things i'm not considering right now is what is culumiative incidence, and whre does it fit in.
