---
title: "Blood Cost modeling"
subtitle: "A brief introduction"
date: "2025-05-16"
author:
  - name: "Harmon Bhasin"
    affiliation: "SecureBio"
format:
  html:
    toc: true
    code-fold: true
    toc-depth: 2
    number-sections: true
    theme: cosmo
  pdf:
    toc: true
---

# Motivation

*I'd like to thank Dan and Simon for our conversations about cost modeling.*

We'd like to estimate the cost of running a blood biosurveillance system in the US for a year that is capable of detecting an HIV like-pathogen at X% cumulative incidence. Specifically, we want to know the sensitivity of blood to detect such a pathogen.

# Background

To get the annual cost to run such a system, one of the main variables we need to estimate is the weekly depth of sequencing. More formally, we want to know the annual sequencing depth required to detect a pathogen at $X$% cumulative incidence with a $r$ weekly growth rate. We simulate this using a stochastic model outlined below. Importantly, we focus on a model where we sequence a pool of samples (where the size is something we vary), as this follows the reality of what getting samples might be like.

## Key Model Assumptions

1. **Exponential Growth**: Infections grow deterministically exponentially to reach cumulative incidence $X$% with growth rate $r$.
2. **Fixed Shedding**: Each infected person sheds for exactly 12 weeks at constant rate (NB: cite study that Lenni bought up during call with Dan)
3. **Random Sampling**: Each week, we sample $P$ people from the total population $N$. This pool represents random sample from total population. Probability of getting sampled is independent of your probability of getting sampled. Additionally, given that $N$ is sufficiently large, we sample with replacement.
4. **Constant mew**: All infected individuals contribute same relative abundance $mew$
5. **Poisson Reads**: Read counts follow Poisson distribution around expected value
6. **Independent Weeks**: Sampling is independent across weeks (with replacement)

## Modeling

### Exponential Growth Model

The number of new infections in week $t$ follows exponential growth:

$$I(t) = I_1 \cdot e^{r(t-1)}$$

where $I_1$ is the initial number of infections in week 1, and $r$ is the weekly growth rate.

**Constraint**: The total cumulative infections must equal the target:

$$\sum_{t=1}^{T} I(t) = \frac{X}{100} \cdot N$$

**Solving for parameters**: Given the exponential sum formula:

$$\sum_{t=1}^{T} I_1 \cdot e^{r(t-1)} = I_1 \cdot \frac{e^{rT} - 1}{e^r - 1} = \frac{X \cdot N}{100}$$

Given the growth rate $r$, cumulative incidence $X$ and initial infection number $I_1$ , we can solve for the time period $T$ in weeks. This gives us an estimate of the number of weeks that it would take for the pathogen to hit $X$ cumulative incidence.

### Shedding Population

The number of people shedding in week t should be those who were infected in the **previous** 12 weeks:

$$S(t) = \sum_{k=\max(1, t-11)}^{t} I(k)$$

This accounts for the 12-week shedding period after infection.

### Sampling Process

Each week, we randomly sample $P$ people from the total population $N$. The number of shedding individuals captured in our sample follows a binomial distribution:

$$\text{Sampled shedders} \sim \text{Binomial}(P, S(t)/N)$$

### Read Detection

The expected number of pathogen reads detected in week $t$ is:

$$E[R(t)] = \text{(sampled shedders)} / P \times mew \times D$$

The actual reads follow a Poisson distribution around this expectation:

$$R(t) \sim \text{Poisson}(\text{(sampled shedders)} / P \times mew \times D)$$

where D is our sequencing depth.

### Detection Criterion

Detection occurs when cumulative reads over all weeks exceed the threshold:

$$R_{cum} = \sum_{t=1}^{T} R(t) \geq \theta$$

### Detection Probability

Due to the stochastic nature of this modeling, it's possible that by time T, we do not hit the read detection threshold, hence, we continue to run the simulation until detection occurs and record the cumulative incidence at that moment.

For each simulation:

1. Run the epidemic week-by-week with stochastic sampling
2. Continue until cumulative reads exceed the threshold: $\sum_{t=1}^{T_{detect}} R(t) \geq \theta$
3. Record the cumulative incidence at detection time $T_{detect}$

This yields a distribution of detection incidences across simulations, from which we can extract:

- $P(\text{detect before X\% incidence})$ for any target X
- Median detection incidence and interquartile range
- Full characterization of early vs late detection scenarios

The optimal sequencing depth $D^*$ per week is then found via binary search to satisfy:
$$D^* = \min\{D : P(\text{detect before X\% incidence}) \geq p_{target}\}$$

## Estimate the relative abundance of an infected individual (mew)


We estimate the expected relative abundance of the HIV like-pathogen for an infected person by using existing metagenomic sequencing data from people with HIV early in their infection. We call this the *mew* ratio.

We look for studies where 1) people were confirmed to be HIV positive, 2) these individual's samples were sequenced separately, 3) they provide the number of HIV reads, along with the total sequencing depth, and 4) they are early in their infection of HIV. We prefer the studies where they provide this information in a table, as it's the fastest to process, and doesn't require us running their data through our pipeline (NB: we will need to highlight the differences between the studies methods if we decide to go with multiple).

## Cost modeling

We estimate our annual cost by breaking the weekly cost into two parts, 1) sequencing based costs, and 2) per-sample collection cost, then summing them over all the weeks in a year:

\begin{align*}
\text{Total cost per week} &= \\
& \underbrace{\text{sequencing depth} \times (\text{cost of sequencing} + \text{cost of processing})}_{\text{Sequencing based costs}} + \\
&\underbrace{\text{cost of single sample} \times \text{number of samples}}_{\text{Per-sample collection cost}} \\
\text{Total cost per year} &= \text{Total cost per week} \times 52
\end{align*}

where:

- sequencing depth: the number of reads required to detect our HIV-like pathogen
- cost of sequencing: the cost of sequencing a pool of samples using an arbitrary sequencing platform
- cost of processing: the cost of sample preparation and library preparation
- cost of single sample: the cost of purchasing and shipping a single sample
- number of samples: the number of samples we get; this will be our pool size (NB: this is an error that I need to fix; right now I assume number of samples is pool size)

# Results

We analyze the results from one study, https://doi.org/10.1128/jvi.00804-18, which "characterized HIV-1 diversity in the blood and female genital tract (FGT) within 2 weeks after detection of infection in three women enrolled in a unique prospective cohort in South Africa." While sequencing reads of HIV for plasma is given, sequnecing dpeth of the sample is given as a range, "40 million to 250 million", so we take the geometric mean to get a read depth of 100 million reads. For the reads, we take the last date from each person.

```{r}
#| label: load-packages
#| output: false

library(tidyverse)
library(furrr)
library(dplyr)

```

```{r}
#| label: aux-functions

# Function to solve for T given r, I1, and X
solve_time_period <- function(r, I1, X_pct, N) {
  target_total <- (X_pct / 100) * N
  if (r == 0) {
    # Constant infections case
    return(target_total / I1)
  } else {
    # Exponential growth case
    # T = (1/r) * ln(1 + (X*N*(e^r - 1))/(100*I1))
    exp_r <- exp(r)
    ratio <- (X_pct * N * (exp_r - 1)) / (100 * I1)
    T <- (1/r) * log(1 + ratio)
    return(T)
  }
}

# Function to calculate infections at time t
calc_infections <- function(t, I1, r) {
  I1 * exp(r * (t - 1))
}

# Function to calculate shedding population at time t
calc_shedding <- function(t, I1, r, shedding_weeks) {
  start_week <- max(1, t - shedding_weeks + 1)
  end_week <- t
  weeks <- start_week:end_week
  infections <- sapply(weeks, function(w) calc_infections(w, I1, r))
  sum(infections)
}

# Function to calculate cumulative incidence as percentage of population
calc_cumulative_incidence <- function(week, I1, r, N) {
  # Calculate total infections from week 1 to target week
  if (r == 0) {
    total_infections <- I1 * week
  } else {
    total_infections <- I1 * (exp(r * week) - 1) / (exp(r) - 1)
  }
  # Return as percentage of population
  return(100 * total_infections / N)
}

```

```{r}
#| label: simulate-detection-distribution

# Function to simulate detection timing distribution
simulate_detection_distribution <- function(r, I1, P, D, theta, mew, shedding_weeks, N, max_weeks = 1000) {
  # Run epidemic week by week until detection
  week <- 1
  total_reads <- 0
  while (week <= max_weeks && total_reads < theta) {
    # Calculate shedding population for current week
    S_t <- calc_shedding(week, I1, r, shedding_weeks)
    S_t <- round(S_t)
    S_t <- min(S_t, N)  # Cap at population size
    if (S_t > 0 && P < N) {
      # For large populations, approximate with binomial
      sampled_shedders <- rbinom(1, P, S_t / N)
      # Generate reads from Poisson distribution
      expected_reads <- sampled_shedders / P * mew * D
      weekly_reads <- rpois(1, expected_reads)
      total_reads <- total_reads + weekly_reads
      if (total_reads >= theta) {
        # Detection occurred!
        cumulative_incidence <- calc_cumulative_incidence(week, I1, r, N)
        return(list(
          detected = TRUE,
          detection_week = week,
          cumulative_incidence = cumulative_incidence,
          total_reads = total_reads
        ))
      }
    }
    week <- week + 1
  }
  # No detection within max_weeks
  return(list(
    detected = FALSE, 
    detection_week = NA, 
    cumulative_incidence = NA,
    total_reads = total_reads
  ))
}

```

```{r}
#| label: analyze-detection-distribution

# Function to analyze detection distribution across many simulations
analyze_detection_distribution <- function(r, I1, P, D, theta, mew, shedding_weeks, N, n_sims = 1000) {
  # Run many simulations
  results <- replicate(n_sims, {
    simulate_detection_distribution(r, I1, P, D, theta, mew, shedding_weeks, N)
  }, simplify = FALSE)
  # Extract detection data
  detected_sims <- results[sapply(results, function(x) x$detected)]
  if (length(detected_sims) == 0) {
    return(list(detection_rate = 0, summary_stats = NA))
  }
  # Extract cumulative incidence at detection
  detection_incidences <- sapply(detected_sims, function(x) x$cumulative_incidence)
  # Calculate summary statistics
  summary_stats <- list(
    detection_rate = length(detected_sims) / n_sims,
    mean_incidence = mean(detection_incidences),
    median_incidence = median(detection_incidences),
    q25_incidence = quantile(detection_incidences, 0.25),
    q75_incidence = quantile(detection_incidences, 0.75),
    min_incidence = min(detection_incidences),
    max_incidence = max(detection_incidences)
  )
  return(list(
    detection_rate = summary_stats$detection_rate,
    summary_stats = summary_stats,
    raw_data = detection_incidences
  ))
}

```

```{r}
#| label: calculate-detection-probability

# Function to calculate detection probability from saved results
calc_detection_probability <- function(detection_results, threshold_incidence) {
  # Calculate P(detect | X% incidence) from saved results
  # detection_results: output from analyze_detection_distribution()
  # threshold_incidence: X% incidence threshold
  if (detection_results$detection_rate == 0) {
    return(0)
  }
  # Count simulations that detected at or before threshold
  detected_early <- sum(detection_results$raw_data <= threshold_incidence)
  total_sims <- length(detection_results$raw_data) / detection_results$detection_rate
  return(detected_early / total_sims)
}

# Function to find optimal sequencing depth using binary search
find_optimal_depth <- function(target_incidence, target_prob = 0.95, r, I1, P, theta, mew, 
                              shedding_weeks, N, n_sims = 500, tolerance = 1e4) {
  cat(sprintf("Finding optimal depth for %.1f%% cumulative incidence at %.0f%% detection probability\n", 
              target_incidence, target_prob * 100))
  # Binary search bounds
  low <- 1e4    # Start with very low depth
  high <- 1e9   # Maximum reasonable depth
  # Track results for visualization
  tested_depths <- c()
  tested_probs <- c()
  while (high - low > tolerance) {
    mid <- round((low + high) / 2)
    cat(sprintf("Testing depth %.2e... ", mid))
    # Run simulations at this depth
    results <- analyze_detection_distribution(r, I1, P, mid, theta, mew, shedding_weeks, N, n_sims)
    p_detect <- calc_detection_probability(results, target_incidence)
    cat(sprintf("P(detect) = %.2f%%\n", p_detect * 100))
    # Store for visualization
    tested_depths <- c(tested_depths, mid)
    tested_probs <- c(tested_probs, p_detect)
    if (p_detect >= target_prob) {
      high <- mid  # Can use less sequencing
    } else {
      low <- mid   # Need more sequencing
    }
  }
  # Run final simulation at selected depth to confirm
  final_results <- analyze_detection_distribution(r, I1, P, high, theta, mew, shedding_weeks, N, n_sims * 2)
  final_p_detect <- calc_detection_probability(final_results, target_incidence)
  cat(sprintf("\nOptimal depth: %.2e reads/week\n", high))
  cat(sprintf("Final P(detect | %.1f%% incidence) = %.2f%%\n", target_incidence, final_p_detect * 100))
  return(list(optimal_depth = high, final_prob = final_p_detect, tested_depths = tested_depths, tested_probs = tested_probs, final_results = final_results))
}

```



```{r}
#| label: setup-functions

find_total_cost <- function(seq_depth, cost_of_seq, cost_of_proc, cost_of_sample, num_samples) {
  return(seq_depth * (cost_of_seq + cost_of_proc) + cost_of_sample * num_samples)
}

```



```{r}
#| label: load-data

# load in data
data <- read_csv("../data/new-hiv-data.csv")

our_data <- data %>% 
  mutate(hiv_ra = hiv_reads / read_depth)

ggplot(our_data, aes(x = hiv_ra)) +
  geom_histogram(binwidth = 0.00002, color = "black", fill = "skyblue") +
  labs(
    title = "Distribution of HIV Relative Abundance (hiv_ra)",
    x = "HIV Relative Abundance (hiv_ra)",
    y = "Count"
  ) +
  theme_minimal()


# Find the mew factor by taking the average of hiv ra over all HIV people.
mew <- our_data %>% summarize(mean = mean(hiv_ra)) %>% pull(mean)

mew <- our_data %>% 
  summarize(geo_mean = exp(mean(log(hiv_ra), na.rm = TRUE))) %>% 
  pull(geo_mean)


```
This study has 3 subjects, who provide us with mew of 5.59e-5, calculated via geometric mean.

Using this, along with the assumption of a population of 300 million people, 100 reads as our detection threshold, a growth rate of 0.01336 which we derived from existing Nature study, initial infection population of 100. We run simulations, where we vary the pool size, and cumulative incidence.


 
```{r}
#| label: run-analysis-temp

# Run analysis to get detection distribution
#cat("Analyzing detection distribution...\n")
#results <- analyze_detection_distribution(r, I1, P, D, theta, mew, shedding_weeks, N, n_sims)
#
### Display summary statistics
#cat("\nDetection Distribution Summary:\n")
#cat(sprintf("Detection rate: %.2f%%\n", results$detection_rate * 100))
#if (results$detection_rate > 0) {
#  cat(sprintf("Mean cumulative incidence at detection: %.2f%%\n", results$summary_stats$mean_incidence))
#  cat(sprintf("Median cumulative incidence at detection: %.2f%%\n", results$summary_stats$median_incidence))
#  cat(sprintf("IQR: %.2f%% - %.2f%%\n", results$summary_stats$q25_incidence, results$summary_stats$q75_incidence))
#  cat(sprintf("Range: %.2f%% - %.2f%%\n", results$summary_stats$min_incidence, results$summary_stats$max_incidence))
#}
#
### Calculate detection probabilities at various cumulative incidence thresholds
#thresholds <- c(1.0, 5.0, 10.0)  # % cumulative incidence
#cat("\n\nDetection Probabilities at Different Cumulative Incidence Levels:\n")
#for (threshold in thresholds) {
#  p_detect <- calc_detection_probability(results, threshold)
#  cat(sprintf("P(detect | %.1f%% cumulative incidence) = %.2f%%\n", threshold, p_detect * 100))
#}
```

```{r}
#| label: find-seq-depth

library(future)
library(future.apply)

## Model parameters
# CONSTANTS
N <- 300e6  # US population
theta <- 100  # detection threshold (reads)
shedding_weeks <- 12
r <- 0.0136      # weekly growth based on HIV
I1 <- 100     # 100 initial infections


# User defined parameters
P <- 1000  # pool size
D <- 1e7   # sequencing depth to test (10 million reads)
n_sims <- 1000  # number of simulations

# Find optimal sequencing depths for different target incidences
cat("\n\nFinding optimal sequencing depths...\n")
target_incidences <- c(1, 5, 10)  # Different cumulative incidence targets
pool_sizes <- c(1000, 3000, 5000)  # Different cumulative incidence targets
optimal_results <- list()


# Parallel
workers <- max(1, parallel::detectCores() - 1)
plan(multisession, workers = workers)

param_grid <- expand.grid(
  target_incidence = target_incidences,
  pool_size       = pool_sizes,
  stringsAsFactors = FALSE
)

#optimal_results <- future_apply(
#  X   = param_grid,          # each row is one job
#  MARGIN = 1,
#  FUN = function(row) {
#    find_optimal_depth(
#      target_incidence = row["target_incidence"],
#      target_prob      = 0.95,
#      r = r, I1 = I1, P = row["pool_size"],
#      theta = theta, mew = mew,
#      shedding_weeks = shedding_weeks,
#      N = N, n_sims = 500
#    )
#  },
#  future.seed = TRUE         # reproducible RNG streams
#)

## After the future_apply() call finishes
#results_dir <- "../results"        # adjust as needed
#if (!dir.exists(results_dir)) dir.create(results_dir)

#saveRDS(optimal_results,
#        file = file.path(results_dir, "optimal_results.rds"))

optimal_results <- readRDS("../results/optimal_results.rds")





#for (i in target_incidences) {
#for (j in pool_sizes) {
#  cat(sprintf("\n--- Target: %.1f%% cumulative incidence ---\n", target_incidences[i]))
#  optimal_results[[i]] <- find_optimal_depth(
#    target_incidence = target_incidences[i],
#    target_prob = 0.95,
#    r = r, I1 = I1, P = j, theta = theta, 
#    mew = mew, shedding_weeks = shedding_weeks, 
#    N = N, n_sims = 500
#  )
#}
#}

optimal_summary <- param_grid %>%                       #  n rows  = length(optimal_results)
  mutate(
    optimal_depth = sapply(optimal_results, `[[`, "optimal_depth"),
    achieved_prob = sapply(optimal_results, `[[`, "final_prob"),
    annual_reads  = optimal_depth * 52
  )

cat("\n\nOptimal Sequencing Depth Summary:\n")
print(optimal_summary)

## Cost analysis for a specific depth
#selected_depth <- 1e7  # 10 million reads
#seq_c <- 2500 / 1e9
#proc_c <- 2500 / 1e9
#
#weekly_cost <- find_total_cost(
#  seq_depth = selected_depth,
#  cost_of_seq = seq_c,
#  cost_of_proc = proc_c,
#  cost_of_sample = 100,
#  num_samples = P
#)
#
#yearly_cost <- weekly_cost * 52
#cat(sprintf("\n\nCost Analysis for D = %.0e reads/week:\n", selected_depth))
#cat(sprintf("Weekly cost: $%.0f\n", weekly_cost))
#cat(sprintf("Annual cost: $%.0f\n", yearly_cost))
```

To calculate the total cost, we use the numbers from the [p2ra paper](https://www.medrxiv.org/content/10.1101/2023.12.22.23300450v4) for the sequencing cost and processing costs, which comes out to be about $2500 each for 1 billion read pairs. For the blood sample costs, I use the higher end of [our estimate](https://docs.google.com/document/d/1rF5fCTDDqa0lZ8Fms9forA-pHMsB2fCcr3dn9ynrW1U/edit?tab=t.0), which puts us at about $100 per sample. Plugging this into our equation with the sequencing depth, we get a cost of about $520K per year.

Hence, given a read threshold of 100, a pool size of 100, and a weekly incidence of 1%, it would cost about 520k per year to detect an HIV-like pathogen.

# Improvements

## Consider the timing of infection in the *mew* estimate

From Lenni:

I think this mostly makes sense but one nuance to consider here is timing of infection. If we want to detect an HIV-like pathogen early, most people will be donating early in the course of infection. HIV has very different shedding dynamics depending on time. Averaging over these states seems mostly reasonable but ideally we'd find people early in the course of infection, asymptomatic, and not taking anti-virals

## Simulating sampling multiple times a week

Right now we simulate sampling once a week. Lenni suggested that we may want to sample more than this (e.g. 2x-3x per week), it would be to figure out what would need to change to allow for this.

## Make some plots

This would be nice to have.

## Adding more studies

We will most likely want to get more data for estimating the mew factor.

## Plot a sweep of different weekly incidences and pool sizes

It would be nice to see how pool size and weekly incidence affect the cost. We could make a plot similar to what is done in the p2ra paper.

## Model prevalence as a function of time

If we wanted to model an outbreak, it would be better to assume a non-constant prevalence.

## Use a longer time horizon for incidence

Right now we use weekly incidence as this is what both the nasal swab and wastewater paper do. We will want to decide whether this is good for us.

## Estimate relative abundance of pool using a distribution

You may have noticed that I look at the median sequencing depth instead of the mean sequencing depth from the simulation. This is because we often have 0 infected people in the pool, however this leads to issues as, we cannot divide by 0. We may decide to model the relative abundance of the pool as a distribution similar to Simon's nasal swab work. 

## Double check all numbers

Double check all the numbers, and make sure they make sense.

# Appendix

## Doubling rate per day for HIV-like pathogen {#doubling-rate-hiv}

For the purpose of epidemiological modeling, a crucial parameter is the **doubling time**—the time it takes for the number of infected individuals to double. To establish a scientifically-grounded value for a new, HIV-like pathogen, we must look to the early phase of the HIV epidemic in the 1970s and 1980s, before widespread awareness and effective treatments altered its natural course.

The standard epidemiological formula for this calculation is:

$$T_d = \frac{\ln(2)}{r}$$

Where:

* `T_d` is the **doubling time** for the year.
* `r` is the **intrinsic growth rate** of the epidemic per year.
* `ln(2)` is the natural logarithm of 2, a constant approximately equal to 0.693.

Luckily, a 2016 study from *Nature*, titled ["1970s and 'Patient 0' HIV-1 genomes illuminate early HIV/AIDS history in North America"](https://www.nature.com/articles/nature19827), found a range for this. The researchers empirically calculated the doubling time (T_d) to be 0.86 -> 1.12 years.

We can take the arithmetic mean of this to get:

$$T_d = 0.99 \text{ years}$$

Converting this to days gives us our final, justifiable figure:

$$0.99 \text{ years} \times 365 \frac{\text{days}}{\text{year}} \approx \textbf{362 days}$$

Based on the most direct empirical evidence from the early HIV epidemic, the justifiable doubling time for a new, unchecked HIV-like pathogen is approximately **362 days**.

For context, this is fundamentally different from a rapidly spreading respiratory virus. Wild-type SARS-CoV-2, for example, had a doubling time of approximately **[3 days](https://naobservatory.org/blog/investigating-the-sensitivity-of-pooled-swab-sampling-for-pathogen-early-detection/)** at the beginning of the pandemic. This stark contrast underscores the slow, persistent transmission dynamic of an HIV-like pathogen compared to the explosive growth of an airborne disease.
