---
title: "Blood Cost modeling"
subtitle: "A brief introduction"
date: "2025-05-16"
author:
  - name: "Harmon Bhasin"
    affiliation: "SecureBio"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    theme: cerulean
---

# Motivation

*A lot of this has been heavily inspired by [Simon's blog post](https://naobservatory.org/blog/investigating-the-sensitivity-of-pooled-swab-sampling-for-pathogen-early-detection/). I thank Simon and Dan for our previous conversations.*


We'd like to make a cost-estimate to run a blood biosurveillance system in the US for a year at a sequencing depth capable of detecting a novel-HIV-like pathogen at X% cumulative incidence.

# Background

Let's first breakdown what exactly it means to detect some pathogen in the population. Overall the process we're interested in can be defined as follows:

- There is the general population, and the virus of interest is circulating in that population with some incidence
- We sample people from this population to get our pools
- For each person in the pool, there is some probability that they're infected with our virus
- We sequence this pool until we hit some number of reads that we determine to be sufficient for detection


Given this framework, we'll an assumption that is similar to Simon's swab post:

- There is some fixed baseline weekly incidence of this virus, we'll assume it to be at 1% (hence our X% cumulative incidence becomes 1% weekly incidence)
  - I use weekly incidence as it allows us to calculate the cost of sequencing by considering the cost in a week, and multiplying it by 52
  - Again, we may decide to do something different later, but I plan on sticking with this for the rest of the analysis.


Given this 1% weekly incidence and the doubling rate of an HIV like-virus to be [0.65](https://bionumbers.hms.harvard.edu/bionumber.aspx?id=105728&ver=6), we can derive prevalence:

```r
  # Inherent to pathogen
  doubling_period_days <- 0.65
  # We set this
  weekly_incidence <- 0.01
  get_prevalence <- function() {
    days_in_week <- 7
    growth_factor <- 2 ^ (1 / doubling_period_days)
    weekly_growth <- growth_factor ^ days_in_week
    prevalence <- weekly_incidence * (weekly_growth) / (weekly_growth - 1)
    return(prevalence)
```

To figure out how deep we need to sequence, we will need to figure out the ratio at which an infected person contributes to HIV relative abundance. We will call this our i2ra factor, and we will estimate this using existing whole blood data coming from people with HIV.

We'd prefer to find studies where 1) confirmed HIV positive people, 2) individual samples, and 3) they provide the number of HIV reads, along with the total sequencing depth. I prefer the studies where they provide these directly in a table, as it's the fastest, and we don't have to justify the use of computational methods (these would be necssary if tried to derive this using the raw sequecning data). In the case that we analyze multiple datasets (which is probably likely), we would want to account for the fact that the datasets use different methods and correct for that, or analyze them separately, similar to how [p2ra does it](https://www.medrxiv.org/content/10.1101/2023.12.22.23300450v4).

To make things simple, I only decided to analyze one study below. However, the study that [I chose](https://naobservatory.org/blog/investigating-the-sensitivity-of-pooled-swab-sampling-for-pathogen-early-detection/) matched the requirements I laid out above pretty well:

- Table 1 gives us practically all the information we need
- All people with HIV, some on [ART](https://hivinfo.nih.gov/understanding-hiv/fact-sheets/hiv-treatment-basics), some not on ART
- Note: I used the standardized viral count for HIV; we probably want to have unstandardized viral count, however for this exploratory work this should be enough
- I looked at a few other studies, but have not included them below.

# Methods

## Variable definitions

- $ps$: pool size
- $i$: infected person
- $i2ra$: infected person to relative abundance
- $n$: sequencing depth
- $d$: detection threshold
- $pr$: prevalence

## Estimating i2ra

To estimate the i2ra, I take our study, and filter the samples to all people not on ART, since it appears to [change the rate of HIV nucleic acid found in blood](https://pmc.ncbi.nlm.nih.gov/articles/PMC4392820/), which may make the HIV read count lower than it actually is. For the individuals that don't have ART, I find the HIV relative abundance by diving the HIV read count by the total read count (before cleaning of course). Using the HIV relative abundance, I take the average over all non-ART individuals. This gives us an estimate of the i2ra factor for an HIV-like pathogen.

$$ i2ra = \sum_{non ART individuals} (HIV reads / total reads) $$

## Estimating sequencing depth

To estimate sequencing depth, we take an approach similar to that of the nasal swab work.

Steps to sequencing depth:
1. Estimate the number of infected people in the pool
2. Calculate the relative abundance of HIV in that pool
3. Use relative abundance along with read threshold to get sequencing depth.

We now discuss each of these in a little bit more detail, but the formula boils down to:

$$ n = \frac{d}{(i/ps) * i2ra}$$

### Estimating the number of infected people in the pool

To get the estimated number of infected people in a pool we might naively multiply the prevalence by the pool size to keep things simple, however, often the prevalence maybe too small to get any people in the pool as infected (e.g. prevalence is 1% but pool size is 50 so 0 people are infected). Hence, what we can do is use a distribution to model people being infected in our pool. Specifically, we can use a Poisson distribution (similar to Simon's work), which nicely accounts for the case where the prevalence may be lower than the pool size. 

$$ n_{infected} \sim Poisson(\lambda=poolsize*prevalence) $$

### Calcualte the relative abundance

To calculate the relative abundance, we can simply take the i2ra factor, and multiply it by the fraction of infected people. THis gives us a point estimate.

$$ ra = (n_{infected}/poolsize) * i2ra $$

### Estimate sequencing depth

To figure out sequencing depth, we can figure out total detection divided by ra to get teh sequencing depth.

$$ sequencingdepth = detectionthreshold / ra $$

## Running simulation

Since we have a non-deterministic portion of our sequencing depth, we can run a simulation to get a distribution of sequencing depth. If we were to switch any of the other estiamtes to be distrbiutions, we would also benefit form this as well. We run simulations so that we can hopefully converge to what the true expected value is rather than just choosing a point estimate.

In this case all that would be changing from estaimte to tesimate is the amount of infected people in the pool.

Going forward then, we can use the mean sequencing depth as the sequencing depth we're interested in.

## Cost estimate

We develop a simple cost model:

Total cost per week = ( mean sequencing depth * (cost of sequencing + cost of processing)) + (cost of single sample * pool_size)


# Analysis


```{r}
#| label: load-packages
#| output: false

library(tidyverse)

```

```{r}
#| label: load-data

# load in data
data <- read_csv("../data/hiv-data.csv")
our_data <- data %>% select("HIV", "HIV load (RNA copies/ml)", "Total no. of reads", "No. of unique reads", "ART")
our_data <- our_data %>% filter(ART == "No") %>% mutate("hiv_ra" = HIV / `Total no. of reads`)
our_data <- our_data %>% rowid_to_column("sample_id")

# Find the i2ra factor by taking the average of hiv ra over all HIV people.
i2ra <- our_data %>% summarize(mean=mean(hiv_ra)) %>% pull(mean)

```

```{r}
#| label: define-functions

find_seq_depth <- function(prevalence, i2ra_factor, pool_size, detection_threshold, simulations) {
  results <- data.frame(
    n_infected = numeric(simulations),
    expected_ra = numeric(simulations),
    required_depth = numeric(simulations)
  )
  for (i in 1:simulations) {
    n_infected <- rpois(1, pool_size * prevalence)
    if (n_infected == 0) {
      expected_ra <- 1e-15
    } else {
      expected_ra <- (n_infected / pool_size) * i2ra_factor
    }
    required_depth <- detection_threshold / expected_ra
    results$n_infected[i] <- n_infected
    results$expected_ra[i] <- expected_ra
    results$required_depth[i] <- required_depth
  }
  summary_stats <- list(
    mean_infected = mean(results$n_infected),
    median_infected = median(results$n_infected),
    zero_pools_pct = 100 * mean(results$n_infected == 0),
    mean_ra = mean(results$expected_ra[results$n_infected > 0]),
    median_ra = median(results$expected_ra[results$n_infected > 0]),
    mean_depth = mean(results$required_depth[is.finite(results$required_depth)]),
    median_depth = median(results$required_depth[is.finite(results$required_depth)]),
    depth_95pct = quantile(results$required_depth[is.finite(results$required_depth)], 0.95)
  )
  return(list(results = results, summary = summary_stats))
}

find_total_cost <- function(seq_depth, cost_of_seq, cost_of_proc, cost_of_sample, num_samples) {
  return(seq_depth * (cost_of_seq + cost_of_proc) + cost_of_sample * num_samples)
}

```


```{r}
#| label: find-seq-depth

static_prevalence <- FALSE
# Choose some parameters
if (static_prevalence) {
  prevalence <- 0.01
} else {
  # Inherent to pathogen
  doubling_period_days <- 0.65
  # We set this
  weekly_incidence <- 0.01
  get_prevalence <- function() {
    days_in_week <- 7
    growth_factor <- 2 ^ (1 / doubling_period_days)
    weekly_growth <- growth_factor ^ days_in_week
    prevalence <- weekly_incidence * (weekly_growth) / (weekly_growth - 1)
    return(prevalence)
  }
  prevalence <- get_prevalence()
}
prevalence

pool_size <- 100
detection_threshold <- 100
number_of_simulations <- 1000

# Get sequencing depth
seq_depth_results <- find_seq_depth(
  prevalence = prevalence,
  i2ra_factor = i2ra,
  pool_size = pool_size,
  detection_threshold = 100,
  simulations = number_of_simulations
)

seq_c <- 2500 / 1e9
proc_c <- 2500 / 1e9
seq_depth <- seq_depth_results$summary$mean_depth

weekly_cost <- find_total_cost(
  seq_depth = seq_depth,
  cost_of_seq = seq_c,
  cost_of_proc = proc_c,
  cost_of_sample = 10,
  num_samples = pool_size
)

yearly_cost <- weekly_cost * 52

#seq_depth <- 8e5
```



# Improvements

## Estimate prevalence not using point

Instead of having a set prevlance, we could use a prevalcne that grows with some time, as they do with the nasal swab work.


This is crucial to being able to relate cumulative incidence to prevalence partically, a simple implemetnation might look like replacing static prevalanec with a time-series function:

I should probably just do this from scratch.

in the nasal swab post, they get weekly prevlance by working with weekly incidence of 1%.


## Estimate relative abundance of pool using a distribution

Instead of taking an average, we could take a more displined approach to modeling the realtive abudnace fro a pool. In the nasal swab work, they fit a log normal distribution to do this.

Right now, we're hurt from infected people who have no viral nucleic acid, we may instead want to have a distribution for this.

We're assuming that the HIV study statistics is represetnative of the whole sample, however this is probably not true.


## Add more studies

We will most likely want to use more studies as this is just one study.


## Use the unstandardized viral count

To some extent, using the standaardized viral counts is going to hurt us.


# Appendix

## Sanity checking cost model

Just for sanity check, we can set the seq_depth to be 8e5, similar to norovirus example in P2ra, paper and we find that we get similar ball park.

Rothman releative abundance for norovirus GII seems to be about 10e-4

> The number of reads required varies substantially between different viruses and studies. For untargeted sequencing, we find that detection at 1% cumulative incidence, with a read threshold of 100 reads, requires 8·7 × 10^5 ... sequencing reads for norovirus GII Rothman.
> At 1% TCI, the predicted annual cost to detect a GII norovirus-like pathogen under Rothman ... was $52,200.

```{r}
#| label: sanity-check

seq_c <- 2500 / 1e9
proc_c <- 2500 / 1e9
seq_depth <- 8e5

weekly_cost <- find_total_cost(
  seq_depth = seq_depth,
  cost_of_seq = seq_c,
  cost_of_proc = proc_c,
  cost_of_sample = 10,
  num_samples = pool_size
)

yearly_cost <- weekly_cost * 52
```
